### 2014-12-15 universalModeTurbulencePaper.tex
Initial commit

%\documentclass{amsart}
%\documentclass[aip,graphicx]{revtex4-1}
\documentclass[12pt,superscriptaddress]{revtex4}
\usepackage{latexsym, color, graphicx}
%\usepackage{latexsym, color, graphicx, comment}
\usepackage{amssymb}
\usepackage{amsmath}

%\graphicspath{{../MATLAB/}}

\newcommand{\red}[1]{\textcolor{red}{#1}}
\newcommand{\vect}[1]{\mbox{\boldmath $#1$}}
\newcommand{\erf}{\mathrm{erf}}
\newcommand{\PS}{Pfirsch-Schl\"{u}ter~}
\newcommand{\ExB}{\vect{E}\times\vect{B}}
\newcommand{\Cnl}{C_{\mathrm{nl}a}}
\newcommand{\vma}{\vect{v}_{\mathrm{m}a}}
\newcommand{\vm}{\vect{v}_{\mathrm{m}}}
\newcommand{\vda}{\vect{v}_{\mathrm{d}a}}
\newcommand{\vE}{\vect{v}_{E}}
\newcommand{\vdao}{\vect{v}_{\mathrm{d}a0}}
\newcommand{\vdo}{\vect{v}_{\mathrm{d}0}}
\newcommand{\vEo}{\vect{v}_{E0}}
\newcommand{\vpar}{v_{||}}
\newcommand{\vperp}{v_{\bot}}
\newcommand{\rhop}{\rho_{\theta}}
\newcommand{\rT}{r_{T}}
\newcommand{\Ti}{T_{\mathrm{i}}}
\newcommand{\rTi}{r_{T\mathrm{i}}}
\newcommand{\rTe}{r_{T\mathrm{e}}}
\newcommand{\reta}{r_{\eta}}
\newcommand{\rn}{r_{n}}
\newcommand{\rperp}{r_\bot}
\newcommand{\fM}{f_{\mathrm{M}}}
\newcommand{\fMa}{f_{\mathrm{M}a}}
\newcommand{\fMi}{f_{\mathrm{M}i}}
\newcommand{\fMe}{f_{\mathrm{Me}}}
\newcommand{\fe}{f_{\mathrm{e}}}
\newcommand{\Te}{T_{\mathrm{e}}}
\newcommand{\vth}{v_{\mathrm{th}}}
\newcommand{\nuii}{\nu_{ii}}
\newcommand{\nuPrime}{\hat{\nu}}
\newcommand{\gradpar}{\nabla_{||}}
\newcommand{\kpar}{k_{||}}
\newcommand{\psia}{\psi_0}
\newcommand{\psiaHat}{\hat{\psi}_0}
\newcommand{\omegase}{\omega_{*}}
\newcommand{\nuee}{\nu_{ee}}
\newcommand{\energy}{\mathcal{E}}
\newcommand{\sgn}{\mathrm{sgn}}
\newcommand{\changed}[1]{#1}
\newcommand{\todo}[1]{\textcolor{red}{#1}}
\newcommand{\transpose}{\mathrm{T}}

\begin{document}

\title{Generalized universal instability: Transient linear amplification and subcritical turbulence}

% repeat the \author .. \affiliation  etc. as needed
% \email, \thanks, \homepage, \altaffiliation all apply to the current author.
% Explanatory text should go in the []'s,
% actual e-mail address or url should go in the {}'s for \email and \homepage.
% Please use the appropriate macro for the type of information

% \affiliation command applies to all authors since the last \affiliation command.
% The \affiliation command should follow the other information.

\author{Matt Landreman}
\email[]{mattland@umd.edu}
\affiliation{Institute for Research in Electronics and Applied Physics, University of Maryland, College Park, MD, 20742, USA}
\author{Gabriel G. Plunk}
\affiliation{Max Planck Institute for Plasma Physics, EURATOM Association and Max-Planck/Princeton Research Center for Plasma Physics}
\author{William Dorland}
\affiliation{Institute for Research in Electronics and Applied Physics, University of Maryland, College Park, MD, 20742, USA}
%\homepage[]{Your web page}
%\thanks{}
%\author{co-authors}

% Collaboration name, if desired (requires use of superscriptaddress option in \documentclass).
% \noaffiliation is required (may also be used with the \author command).
%\collaboration{}
%\noaffiliation

\date{\today}

\begin{abstract}

In this work we numerically demonstrate both significant transient (i.e. non-modal) linear amplification
and sustained nonlinear turbulence
in a kinetic system that is linearly stable.
The particular system considered is an electrostatic slab
with magnetic shear, kinetic electrons and ions,
weak collisions,
and a density gradient, but with no temperature gradient.
In contrast to hydrodynamic examples of non-modal growth and subcritical turbulence,
here there is not sheared flow in the equilibrium.
Significant transient linear amplification
is found when the magnetic shear and collisionality are weak.
It is also demonstrated that nonlinear turbulence can be sustained if initialized
at sufficient amplitude.
We prove these two phenomena are related:
when sustained turbulence occurs without instability,
states that are typical of the turbulence must
yield transient linear amplification of the gyrokinetic free energy.

\end{abstract}

\pacs{}% insert suggested PACS numbers in braces on next line

\maketitle %\maketitle must follow title, authors, abstract and \pacs

\section{Introduction}

Despite the usefulness of linear eigenmode analysis in many contexts,
it has been established that
such analysis can in practice give misleading predictions
about the stability of various systems \cite{TrefethenEmbree}.
In particular, turbulence may be sustained in a number of fluid systems
despite the absence of any unstable eigenmodes.
The first such systems known were certain sheared flow patterns in
neutral fluids, such as plane Poiseuille or plane Couette flow
\cite{Tillmark,TrefethenSubcritical,Grossman}.
Later, sustained turbulence was seen
in fluid simulations of several plasma systems with no absolute linear instabilities
\cite{Waltz, Scott1, Scott2, Drake}.
More recently,
turbulence in the absence of linear instability has been seen in kinetic plasma simulations involving
sheared equilibrium flow \cite{EdmundPRL, BarnesRotation}.
These examples demonstrate that linear eigenmode analysis can give
deceiving predictions for nonlinear behavior,
a conclusion supported by other studies of plasma turbulence in which
linear instability exists, but
the fluctuations in saturated nonlinear turbulence are different in character from the linear
eigenmodes \cite{Biskamp,Friedman1,Friedman2}.

For the cases of turbulence without absolute linear stability in the presence of sheared flows,
for both neutral fluids and plasmas, the phenomenon of transient linear amplification
appears to be important
\cite{TrefethenSubcritical, FarrellIoannou, Grossman, DelSoleSurvey, TrefethenEmbree, Schmid, Newton, Alex}.
Also known as non-modal amplification, transient amplification is a period of growth before
the exponential decay sets in.
A necessary (but not sufficient) condition for transient amplification
is that the linear operator must not have a complete orthogonal set of eigenvectors.
Analysis of transient amplification has also been applied to plasma systems in which
linear instability exists but transient linear behavior may be
more important \cite{Camargo, Camporeale2010, Friedman2, SquirePRL, SquireApJ}.

In the present work, we consider the phenomena of transient amplification and turbulence
without instability for the case of the weakly-collisional
plasma drift wave driven by a density gradient\cite{Galeev,Krall3},
neglecting toroidal effects.
This oscillation has been called the ``universal instability'' as it was at first
mistakenly thought to be linearly unstable in any plasma with a density gradient.
The oscillation is electrostatic in nature, and linear instability can indeed be found
in slab geometry without magnetic shear or temperature gradients.
It was later established that when any amount of magnetic shear is included in the model,
no matter how small,
the eigenmodes become linearly damped for $k_y \rho_i \ll 1$
\cite{Ross, Tsang, Antonsen}.
However, instability can persist at $k_y \rho_i \ge 1$ when the collisionality is
very low \cite{usUniversalInstability}.
When both magnetic shear and modest collisions are included \cite{usUniversalInstability},
the mode becomes linearly stable at all $k_y \rho_i$.
It is this latter situation that we will consider here.

The fluid analogue of this drift wave system was investigated in \cite{Drake},
using a model with three fluctuating quantities (density, electrostatic potential, and parallel velocity).
In this model all linear modes were damped, yet nonlinearly, turbulence could be sustained.
It was pointed out that an important secondary
instability in this system is a drift wave rotated by $\pi/2$ in the plane
perpendicular to the magnetic field, since the stabilizing effects of magnetic
shear are reduced in this orientation.
In this rotated drift instability, the role of the equilibrium density gradient is replaced in a sense by the
perturbed density gradient, the latter represented in the nonlinear term.
However, since the nonlinearity is conservative, the rotated drift wave does not inject energy
into the system in the way that can be done by a perturbation feeding off the density gradient of the equilibrium.
Several authors have also investigated transient linear effects in fluid drift wave models
\cite{Camargo, Friedman1, Friedman2}. However, these studies did not
include magnetic shear, which we will find to have a very strong effect on
transient behavior.

In the following sections, we establish three results concerning the drift wave system in slab geometry
with magnetic shear,
extending previous work by considering a kinetic description and exploring the role of transient linear amplification.
For these studies, the initial-value gyrokinetic code gs2 \cite{gs2} will be used
for numerical computations.
First, in section \ref{sec:transient} we demonstrate that
significant transient linear amplification is possible in this system,
as long as the magnetic shear is not too strong.
Second, in section \ref{sec:turbulence} we demonstrate that universal-mode turbulence without linear
instability, seen previously in the fluid model in \cite{Drake},
is also possible in the gyrokinetic model.
Finally, we prove there is a relationship between
transient linear amplification and nonlinear turbulence in this system,
using similar arguments to the proof for hydrodynamics in \cite{DelSoleNecessity}
and arriving at similar conclusions.
Specifically, if the physical parameters are such that turbulence can be sustained,
typical states of the turbulence must be amplified linearly for some amount of time, even if these states
eventually decay exponentially.
Two more precise statements of this result and their proof will be given in section \ref{sec:proof}.

\section{Gyrokinetic model}
\label{sec:model}

We consider slab geometry with magnetic shear, so the magnetic field in Cartesian
coordinates $(x',y',z')$ is $\vect{B} = [\vect{e}_{z'} + (x' / L_s) \vect{e}_{y'} ]B$
for some constant $B$ and shear length $L_s$.  A field-aligned coordinate system
is introduced: $x = x', \;y = y' - x' z' / L_s, \; z = z'$,
so $\vect{B}\cdot\nabla x = \vect{B}\cdot\nabla y = 0$ and $\vect{B}\cdot\nabla z = B$.
In this geometry, we consider the nonlinear electrostatic
gyrokinetic equation \cite{FriemanChen},
dropping toroidal effects:
\begin{eqnarray}
\label{eq:gke}
\frac{\partial h_s}{\partial t}
+ v_{||} \frac{\partial h_s}{\partial z}
+ \frac{1}{B} \left[
\frac{\partial \left<\Phi\right>_{\vect{R}_s}}{\partial X_s} \frac{\partial h_s}{\partial Y_s}
-\frac{\partial \left<\Phi\right>_{\vect{R}_s}}{\partial Y_s} \frac{\partial h_s}{\partial X_s}
\right]
-\left< C_s\left\{h_s\right\} \right>_{\vect{R}_s} \\
=
\frac{q_s}{T_s} f_{Ms} \frac{\partial \left<\Phi\right>_{\vect{R}_s}}{\partial t}
-\frac{f_{Ms}}{B L_n} \frac{\partial \left< \Phi\right>_{\vect{R}_s}}{\partial Y_s}
\left[ 1 + \left( \energy-\frac{3}{2}\right) \eta_s\right]
\nonumber
\end{eqnarray}
together with the quasineutrality condition
\begin{equation}
-\frac{n e \Phi}{T_i} + \int d^3v \left<h_i\right>_{\vect{r}}
=
\frac{n e \Phi}{T_e} + \int d^3v \left<h_e\right>_{\vect{r}}.
\label{eq:qn}
\end{equation}
Here $s \in \left\{i,e\right\}$ subscripts denote species,
$\Phi(t,\vect{r})$ is the electrostatic potential, $h_s(t,\vect{R}_s,\energy,\mu)$ is the nonadiabatic distribution function,
$\energy = v^2/2$, $\mu = v_{\bot}^2/(2B)$,
and $f_{Ms}$ is the leading-order Maxwellian.
The scale length of the equilibrium density $n$ is $L_n = -(1/n)(dn/dx)$,
$q_i = e = -q_e$ is the proton charge,
$T_s$ is the species temperature,
$C_s$ is the collision operator,
and
$\left<\right>_{\vect{R}_s}$ and $\left<\right>_{\vect{r}}$ denote
gyroaverages at fixed particle position $\vect{r}=(x,y,z)$ and guiding-center position $\vect{R}_s=(X_s,Y_s,Z_s)$.
The distinction between $z$ and $Z_s$ may be neglected.
Expanding the fluctuating quantities in spatial Fourier modes as $\Phi = \sum_{k_x,k_y} \Phi_{k_x,k_y}(t,z) \exp(i k_x x + i k_y y)$
and $h_s = \sum_{k_x,k_y} h_{s,k_x,k_y}(t,z,\energy,\mu) \exp(i k_x X_s + i k_y Y_s)$,
we may replace the gyro-averages with Bessel functions:
$\left<\Phi_{k_x,k_y} \exp(i k_x x + i k_y y)\right>_{\vect{R}_s}
=J_{0s} \Phi_{k_x,k_y} \exp(i k_x X_s + i k_y Y_s)$
and
$\left<h_{s,k_x,k_y}\exp(i k_x X_s + i k_y Y_s)\right>_{\vect{r}} = J_{0s} h_{s,k_x,k_y}\exp(i k_x x + i k_y y)$
with $J_{0s} = J_0(k_{\perp} v_{\perp} / \Omega_s)$
and $\Omega_s = e B/m_s$.
The perpendicular wavenumber is given by
\begin{equation}
k_{\perp} = \sqrt{k_y^2 + (k_x - k_y z/L_s)^2}.
\label{eq:kperp}
\end{equation}
The system is uniquely specified by the following dimensionless parameters:
$L_s / L_n$, $\eta_i$, $\eta_e$, $T_e / T_i$, and $k_y \rho_i \nu_{ee} / \omega_*$,
where $\nu_{ee} = \sqrt{2}\pi n e^4 \ln\Lambda/[(4\pi\epsilon_0)^2 m_e^{1/2} T_e^{3/2}]$ is the electron-electron collision frequency,
and $\omega_* = k_y T_e/(e B L_n)$ is the electron drift frequency.

Notice that the parallel coordinate $z$ enters the system (\ref{eq:gke})-(\ref{eq:kperp}) only through (\ref{eq:kperp}).
Also notice $k_x$ enters the system (\ref{eq:gke})-(\ref{eq:kperp}) only through (\ref{eq:kperp}) and through the nonlinear term in (\ref{eq:gke}).
Therefore, a linear mode at nonzero $k_x$ and nonzero $k_y$ satisfies exactly the same equations as the corresponding $k_x=0$ mode, but translated in $z$ by
$L_s k_x / k_y$.  As a consequence, the linear growth or damping rates of $k_y \ne 0$ modes
are independent of $k_x$, and the eigenmode structures are independent of $k_x$
up to this translation in $z$.

One important property of the system (\ref{eq:gke})-(\ref{eq:qn})
is the free energy conservation equation \cite{Howes, AbelReview}
obtained from the $\sum_s \int d^3R_s \int d^3v (T_s h_s/f_{Ms}) (\ldots)$ moment of (\ref{eq:gke}):
\begin{equation}
\frac{d W}{d t} = S(\Phi,h_i,h_e)
\label{eq:dWdt}
\end{equation}
where
\begin{equation}
\label{eq:W}
W = \sum_s \int d^3r \int d^3v \frac{T_s \, \delta\! f_s^2}{2 f_{Ms}}
\end{equation}
is the free energy, $\delta\! f_s = h_s - q_s \Phi f_{Ms}/T_s$ is the total departure of the distribution
function from a Maxwellian,
and
\begin{eqnarray}
\label{eq:freeEnergySources}
S(\Phi,h_i,h_e)
&=&
\sum_s \int d^3r \int d^3v \frac{T_s h_s}{f_{Ms}}  C_s\{ h_s \}  \\
&&+\sum_s
\int d^3R_s \int d^3v \frac{h_s}{B} \frac{\partial \left< \Phi\right>_{\vect{R}_s}}{\partial Y_s}
 \frac{T_s}{n_s}\frac{dn_s}{dx}
\left[ 1 + \left( \energy-\frac{3}{2}\right) \eta_s\right]
\nonumber
\end{eqnarray}
is the net energy source.
(To derive (\ref{eq:dWdt})-(\ref{eq:freeEnergySources}), the equalities
$\int d^3 R_s \int d^3v \left< \ldots\right>_{\vect{R}_s}
= \int d^3 R_s \int d^3v (\ldots)
\approx \int d^3 r \int d^3v (\ldots)
= \int d^3 r \int d^3v \left< \ldots\right>_{\vect{r}}
$ are used.)
The first term on the right-hand side of (\ref{eq:freeEnergySources})
is negative-definite (by the $H$-theorem).
The second term has the form of the particle and heat fluxes multiplied by
the density and temperature gradients, and this term increases $W$ if the fluxes have the appropriate sign
to relax the gradients.
Thus, in a turbulent steady state where $W$ is statistically constant,
equations (\ref{eq:dWdt})-(\ref{eq:freeEnergySources}) indicate
that energy is injected by the equilibrium gradient(s)
and dissipated by collisions.
Note that the operation $\int d^3R_s$ used in the derivation of
(\ref{eq:dWdt}) annihilates the nonlinear term
in (\ref{eq:gke}), and so (\ref{eq:dWdt})-(\ref{eq:freeEnergySources})
apply to both the linear and nonlinear dynamics.

The system (\ref{eq:gke})-(\ref{eq:qn}) may or may not possess absolute linear instability,
depending on the physical parameters.
For the rest of this paper, we take $\eta_i = \eta_e = 0$
so it is certain that the ion-temperature-gradient (ITG) and electron-temperature-gradient (ETG)
instabilities are suppressed.
Even without temperature gradients, the system may still be unstable to the universal instability
driven by the density gradient.
For reference, the growth rates and frequencies of this instability
in the absence of magnetic shear are given in Appendix \ref{appendix:zeroShear}.
As this instability is stabilized for $k_y \rho_i \ll 1$ when even a small amount of magnetic shear
is introduced \cite{Ross, Tsang}, we might expect that this system can exhibit transient linear amplification, with an amount of
amplification that tends to infinity as the magnetic shear tends to zero.  In this way, the apparently singular limit
of zero shear may be seen as a continuous limit of the general shear case.

\section{Transient linear amplification}
\label{sec:transient}

Turning now to linear initial-value simulations,
we indeed find that significant transient linear amplification is possible for certain parameters.
Figure \ref{fig:transientAmplification}.a illustrates
a linear computation for a single $(k_x,k_y)$ in a case of very weak magnetic shear, $L_s / L_n = 300$.
Other parameters used are $k_y \rho_i = 1$, $\nu_{ee} / \omega_* = 0.1$,
$m_i = 3672 m_e$,
$T_e = T_i$, and $\eta_i = \eta_e = 0$;
the results are independent of $k_x$ as noted previously.
Collisions are also included to eliminate absolute instability for $k_y \rho_i \ge 0.7$,
as discussed in \cite{usUniversalInstability},
using the collision operators detailed in Ref. \cite{Abel}.
As transient amplification generally depends on the choice
of norm, here we compare three reasonable (squared) norms: $W$, $\int dz |\Phi|^2$,
and the maximum of $|\Phi|^2$ along $z$.
The curve for each norm is scaled by a constant so the initial amplitude is 1.
Amplification by several orders of magnitude is observed in all norms
before exponential decay sets in.

Results for the three norms shown are quite similar,
though the curve for the $W$ norm appears slightly higher than the other curves
due to less decay in the initial period $t v_i / L_n <20$.
Indeed, we find
all results described or plotted for the remainder of this paper are quite similar
for all of these norms.
Henceforth we restrict our attention to the $W$ norm since it is well motivated
by the conservation equation (\ref{eq:dWdt}).

For the calculation in figure \ref{fig:transientAmplification}, the initial condition used
is the standard ``noise'' initial condition of gs2:
$g_s = h_s - \left< \Phi\right>_{\vect{R}_s} q_s f_{Ms} / T_s$ is set to a Maxwellian velocity distribution
times a random complex number at each grid point in $z$.
(The real and imaginary parts of these complex numbers are each taken from a uniform distribution on $[-1,1]$.)
While techniques exist to calculate the optimal perturbation
that gives maximum amplification \cite{FarrellIoannou, SquirePRL, SquireApJ}, such techniques
are numerically challenging for this kinetic model due to the high dimensionality
of the problem.
The noise initial condition is a convenient shortcut, ensuring energy is likely
to be given to a growing perturbation if one exists.
It has been shown in \cite{Camargo, SquireApJ} that for studies of transient linear
amplification, random initial conditions are a reasonable proxy
for a directly calculated optimal initial condition.

\begin{figure}[h!]
% Figure generated with m20141119_01_plotGS2TransientAmplificationForPaper_figureA.m
\includegraphics[width=3in]{m20141119_01_plotGS2TransientAmplificationForPaper_figureA_a.pdf}
% Figure generated with m20141121_01_plotGS2TransientAmplificationForPaper.m
\includegraphics[width=3in]{m20141121_01_plotGS2TransientAmplificationForPaper_figuresBC_b.pdf}
\includegraphics[width=3in]{m20141121_01_plotGS2TransientAmplificationForPaper_figuresBC_c.pdf}
\caption{(Color online) (a) Linear evolution for $L_s/L_n=300$, $\eta_i = \eta_e = 0$,
$\nu_{ee} = 0.1 \omega_*$, and $k_y \rho_i = 1$, illustrating transient amplification in three norms.
The initial conditions are detailed in the main text.
(b) The amount of transient amplification may be defined relative to $t=0$ or relative
to the minimum that occurs shortly thereafter.
(c) The potential structure along a field line for this example.
\label{fig:transientAmplification}}
\end{figure}

Figure \ref{fig:transientAmplification}.b shows the same computation, zooming in to the initial transient period.
Before the amplification begins around $t v_i/L_n=20$, there is an initial decay period due to strongly
damped modes. An amplification factor may be defined relative to either to this minimum amplitude,
or relative to the amplitude at $t=0$.
In both cases, we define the amplification factor to be 1
if there is no such transient amplification.

Transient amplification is exponentially reduced by increasing magnetic shear,
as illustrated in figure \ref{fig:transientAmplificationVsLsOverLn}.
For this figure, $k_y \rho_i$ is fixed at 2, and other parameters
are the same as in figure \ref{fig:transientAmplification}.
(We choose $k_y \rho_i=2$ here since
wavenumbers near this value will turn out to be the most
amplified, as we will find shortly.)
There is some scatter in figure \ref{fig:transientAmplificationVsLsOverLn}, associated with the random initial condition,
but the exponential trend with $L_s/L_n$ is clear.
Notice that amplification only appears possible if $L_s/L_n \gtrsim 20$.
Values of $L_s/L_n$ far in excess of 20 are obtained in the edge of tokamak experiments.
(For a high aspect ratio tokamak, $L_s \sim qR/\hat{s}$ where $q$ is the safety factor, $R$ is
the major radius, $\hat{s} = (r/q) dq/dr$, and $r$ is the minor radius.)
Figure \ref{fig:transientAmplificationVsLsOverLn}.b demonstrates that the time during which amplification occurs
varies approximately linearly with $L_s/L_n$.

Figure \ref{fig:transientAmplificationVsKy} shows the dependence
of the amplification factor on $k_y$ at fixed $L_s/L_n = 200$.
Significant amplification can be observed for a range of $k_y$ surrounding
$1/\rho_i$.
Results are shown for two values of collisionality,
$\nuee k_y \rho_i / \omega_*= 0.2$ and 1.
(When the collisionality is much smaller than these values,
absolute linear instability is present \cite{usUniversalInstability}.)
It can be seen that increasing collisionality tends to limit amplification
at larger $k_y$ while having much less effect at smaller $k_y$.
This kinetic problem is similar to conventional hydrodynamics in that for both cases, non-modal amplification
decreases with increasing collisional dissipation (associated with viscosity in the Navier-Stokes equation
for the hydrodynamic case.)
However, the kinetic problem here is rather different
from the hydrodynamic case in the sensitivity of the transient amplification
to the geometric parameter $L_s/L_n$,
which is not associated with entropy-producing collisions.

\begin{figure}[h!]
% Data computed on Edison in /global/scratch2/sd/landrema/20141121-01-coderunner-edison-transientAmplificationScans
% Plot generated by first running
% m20141121_02_plotTransientAmplificationLsOverLnScanForPaper.m on edison,
% and then running
% m20140723_03_plotTransientAmplificationLsOverLnScanForPaper.m on my laptop.
% ----------------------------------------
% Data for previous version of figure computed on Edison in /global/scratch2/sd/landrema/20140718-01-coderunner-edison-transientAmplificationScans
% Plot generated by first running
% m20140725_02_plotTransientAmplificationLsOverLnScanForPaper.m on edison,
% and then running
% m20140723_03_plotTransientAmplificationLsOverLnScanForPaper.m on my laptop.
\includegraphics[width=3in]{m20141121_03_plotTransientAmplificationLsOverLnScanForPaper_a.pdf}
\includegraphics[width=3in]{m20141121_03_plotTransientAmplificationLsOverLnScanForPaper_b.pdf}
%\includegraphics[width=3in]{m20140723_03_plotTransientAmplificationLsOverLnScanForPaper.pdf}
\caption{(Color online) The (a) number of $e$-foldings of transient amplification and (b)
duration of the amplification both scale with the magnetic shear length scale $L_s$,
consistent with the appearance of absolute instability in the limit of vanishing shear ($L_s \to \infty$).
Parameters used are $k_y \rho_i = 2$ and $\nu_{ee} = 0.1 \omega_*/(k_y \rho_i)$.
\label{fig:transientAmplificationVsLsOverLn}}
\end{figure}

\begin{figure}[h!]
% Data computed on Edison in /global/scratch2/sd/landrema/20141121-02-coderunner-edison-transientAmplificationKyScans
% Plot generated by first running
% m20140725_01_plotTransientAmplificationKyScanForPaper.m on edison,
% and then running
% m20140728_01_plotTransientAmplificationKyScans.m on my laptop.
% ----------------------------------------
% Data for previous version of figure computed on Edison in /global/scratch2/sd/landrema/20140729-01-coderunner-edison-transientAmplificationKyScans
% Plot generated by first running
% m20140725_01_plotTransientAmplificationKyScanForPaper.m on edison,
% and then running
% m20140728_01_plotTransientAmplificationKyScans.m on my laptop.
\includegraphics[width=3in]{m20141122_01_plotTransientAmplificationKyScans.pdf}
%\includegraphics[width=3in]{m20140728_01_plotTransientAmplificationKyScans.pdf}
\caption{(Color online) Transient amplification can be observed
for a range of $k_y$ surrounding $1/\rho_i$.  For this figure, $L_s/L_n=200$.
The two collisionalities shown are $\nuee k_y \rho_i/\omega_* = 0.2$ and 1.
\label{fig:transientAmplificationVsKy}}
\end{figure}

\section{Turbulence}
\label{sec:turbulence}

We now demonstrate that when the nonlinear term in (\ref{eq:gke})
is retained, sustained turbulence is possible
in this system.
For the demonstration here we consider
the parameters $L_s/L_n = 20$, $T_e = T_i$, $m_i = 3672 m_e$, $\eta_i = \eta_e = 0$, and $\nu_{ee} k_y \rho_i / \omegase = 0.1$.
We find this point in parameter-space is approximately on the edge of where sustained turbulence
is possible: if the magnetic shear or collisionality is increased much beyond these values,
turbulence cannot be sustained indefinitely (though it may persist for a finite time.)
Note that $L_s/L_n=20$ is also approximately the maximum shear
in figure \ref{fig:transientAmplificationVsLsOverLn} for which transient linear amplification is observed
for random initial conditions.
(While absolute linear instability was found in \cite{usUniversalInstability} for $L_s/L_n \ge 17$ in the limit of vanishing collisionality, here the finite collisionality is enough to suppress this linear instability.)
Before presenting nonlinear simulations,
we first illustrate the linear behavior of the system using all the physical and numerical
parameters that will be used for the nonlinear runs.  Figure \ref{fig:linearBox} shows the time evolution of
$W$ for all the dealiased $k_y$ modes that will be
used for nonlinear simulations, including twist-and-shift boundary conditions,
but with the nonlinearity turned off, and
using the random initial conditions described in section \ref{sec:transient}.
More precisely, the quantities plotted are $W_{k_x,k_y}$, summed over the $k_x$ that link to $k_x=0$, where
\begin{equation}
\label{eq:Wk}
W_{k_x,k_y} = L_x L_y \sum_s \int dz \int d^3v \frac{T_s}{2 f_{Ms}}\left|\delta\! f_{s,k_x,k_y}\right|^2
\end{equation}
are the 2D Fourier contributions to $W = \sum_{k_x,k_y} W_{k_x,k_y}$.
Here,
$L_x$ and $L_y$ are the sizes of the simulation domain in $x$ and $y$,
and $\delta\! f_s =\sum_{k_x,k_y} \delta\! f_{s,k_x,k_y} \exp( i k_x x+i k_y y)$.
Curves for the other $k_x$ in the simulation look similar to those in figure \ref{fig:linearBox}.
The amplitude of each
of these curves decreases for large $t$, confirming there is no instability for these physical and
numerical parameters.  The decrease is mostly monotonic, but there is typically a small
amount of transient growth by a factor $ \lesssim 2\times$ for $k_y \rho_i$ between 1-2.
Thus, this point in parameter space is on the edge of the region in which transient linear
amplification occurs, as anticipated from figure \ref{fig:transientAmplificationVsLsOverLn}.

Throughout this section, the numerical parameters used were as follows: $12 \times 2$ grid points in pitch angle and $\sgn(v_{||})$, 12 grid points in energy, 24 grid points per cell in $z$ (with cells at $k_y=j \Delta k_y$ linked if separated in $k_x$ by $4 j \Delta k_x$), CFL number 0.5, $256 \times 64$ modes in $x$ and $y$ before dealiasing, and box size $(L_x, L_y, L_z) = (31.4 \rho_i, \; 62.8 \rho_i, \; 160 L_n)$.  It was verified that both the linear and nonlinear results reported in this section were not significantly altered under factor-of-2 changes in any of these numerical parameters.

One noteworthy linear property apparent in figure \ref{fig:linearBox}
is that damping of modes with $k_y=0$ is very weak,
nearly invisible on the scale of the figure.
Linear modes with $k_y=0$ and $\partial/\partial z=0$ are damped only by collisions.
(In the absence of any collisions, any $h_e$ and $h_i$ that depend only on $\energy$ and $\mu$ are steady-state solutions of the linearized system.)
Collisions introduce damping of these $k_y=0$ modes, with a damping rate that increases with $k_x^2$ due to the classical transport terms in the gyroaveraged collision operator.

\begin{figure}[h!]
% Dataset is /global/scratch2/sd/landrema/20141112-01-coderunner-edison-subcritical-box-lowLsOverLn/v/id_59
% Figure was generated by first running
% /global/u2/l/landrema/MATLAB/m20141124_01_processGS2_linearBoxFigureForPaper.m on edison,
% then running
% m20141126_01_plotWVsTimeForPaper_linearBox.m on my laptop
% -------------------------------------
% Previous version of figure:
% Dataset is /global/scratch2/sd/landrema/20140730-01-coderunner-edison-subcritical-box/v/id_89
% Figure was generated by first running
% /global/u2/l/landrema/MATLAB/m20140804_02_extractPhi2ByKyFromGS2_linearBox.m on edison,
% then running
% m20140804_03_plotPhi2VsTimeForPaper_linearBox.m on my laptop
\includegraphics[width=6.5in]{m20141126_01_plotWVsTimeForPaper_linearBox.pdf}
%\includegraphics[width=6.5in]{m20140804_03_plotPhi2VsTimeForPaper_linearBox.pdf}
\caption{(Color online)
Linear dynamics for the parameters used in the
nonlinear simulations of figure \ref{fig:amplitudeDependence}.
The large-$t$ behavior demonstrates there is no absolute linear instability.
A small amount of transient amplification can be seen.
\label{fig:linearBox}}
\end{figure}

Finally, figure \ref{fig:amplitudeDependence} shows two nonlinear simulations for the same parameters as figure \ref{fig:linearBox},
differing only in the initial amplitude of the $k_y=0.1$ part of the perturbation.
For sufficiently large initial amplitude, turbulence is sustained,
demonstrating nonlinear instability.

A variety of strategies can be used to initiate the nonlinear instability in numerical simulations.
Large-amplitude noise can be used, but this method is numerically inconvenient (at least in gs2) due to the rapid
drop in amplitude over the first few timesteps due to strongly damped modes,
which causes significant changes in the CFL-stable timestep associated
with the perpendicular $\vect{E}\times\vect{B}$ speed,
thereby requiring recomputation of the implicit time-advance operator.
Another method for initiating the nonlinear instability is to begin with a sufficient
temperature gradient to bring about linear instability, and then to reduce the temperature gradient
once the turbulence reaches a sufficient amplitude \cite{Drake}.
A third effective method, the one used in figure \ref{fig:amplitudeDependence},
is to initialize with one of the least-damped linear eigenmodes scaled up to appropriate amplitude, with noise in the other modes.  We find this method to be particularly convenient since variation in the stable timestep
is minimized.

\begin{figure}[h!]
% Datasets are in /global/scratch2/sd/landrema/20141112-01-coderunner-edison-subcritical-box-lowLsOverLn
% runs 54,55,58
% Figure is produced by first running
% m20141125_01_processGS2_amplitudeDependence.m on edison,
% then running
% m20141125_02_WVsTimeForPaper_amplitudeDependence.m on my laptop.
% ---------------------------------------
% Old version of figure:
% Datasets are in /global/scratch2/sd/landrema/20140730-01-coderunner-edison-subcritical-box
% runs 93,97,95.
% Figure is produced by first running
% m20140804_03_extractPhi2ByKyFromGS2_amplitudeDependence.m on edison,
% then running
% m20140804_05_plotPhi2VsTimeForPaper_amplitudeDependence.m on my laptop.
\includegraphics[width=6.5in]{m20141125_02_WVsTimeForPaper_amplitudeDependence.pdf}
%\includegraphics[width=6.5in]{m20140804_05_plotPhi2VsTimeForPaper_amplitudeDependence.pdf}
\caption{(Color online)
Simulations for exactly the same physical and numerical parameters as the decaying
linear simulation in figure (\ref{fig:linearBox}), but now including nonlinearity and varying the
initial amplitude of the $k_y=0.1$ component.
(a) When the initial amplitude is sufficient, turbulence grows and is sustained, whereas (b) for small initial amplitude,
all modes decay as in a linear simulation.
\label{fig:amplitudeDependence}}
\end{figure}

As another demonstration that the system displays nonlinear instability without linear instability,
figure \ref{fig:linearRestart} shows the result of
turning off the nonlinear term after a period of saturated turbulence.
The nonlinear term is turned off at time $t v_i / L_n = 2064$, shown as a vertical
dotted line.  All modes eventually decay.
However some modes
display a modest amount ($\lesssim 3\times$) of transient amplification before decaying.

\begin{figure}[h!]
% Simulations in /global/scratch2/sd/landrema/20141112-01-coderunner-edison-subcritical-box-lowLsOverLn
% Runs: 58 (nonlinear) and 61 (linear restart)
% Figure generated by first running
% m20141126_01_processGS2_linearRestartFigureForPaper.m on edison,
% then running
% m20141126_02_plotWVsTimeForPaper_linearRestart.m on my laptop.
% -------------------------------
% Previous version of figure:
% Simulations in /global/scratch2/sd/landrema/20140730-01-coderunner-edison-subcritical-box
% Figure generated by first running
% m20140806_01_extractPhi2ByKyFromGS2_linearRestart.m on edison,
% then running
% m20140807_02_plotPhi2VsTimeForPaper_linearRestart.m on my laptop.
\includegraphics[width=6.5in]{m20141126_02_plotWVsTimeForPaper_linearRestart.pdf}
%\includegraphics[width=6.5in]{m20140807_02_plotPhi2VsTimeForPaper_linearRestart.pdf}
\caption{(Color online)
Demonstration of the concepts in section \ref{sec:proof}.
Curves show free energy in the 2D modes with $k_x=0$ (including intervals linked by
the twist-and-shift parallel boundary condition) for various $k_y$.
After a period of saturated turbulence, the nonlinear term is turned off at normalized
time 2064, denoted by the vertical dashed line.
All modes eventually decay, but transient amplification is apparent first
for some of the 2D modes. Such amplification must occur in at least one 2D mode in such a linear ``restart'' from
steady turbulence, as proved in section \ref{sec:proof}.
\label{fig:linearRestart}}
\end{figure}

The particle and heat fluxes for a long nonlinear simulation at these parameters are plotted in figure \ref{fig:fluxes}.
As the time average of $S$ in (\ref{eq:dWdt})-(\ref{eq:freeEnergySources}) must vanish, and since the collision term
is negative-definite, there must be an average particle flux in the appropriate direction
to relax the equilibrium density gradient, and indeed we find such a flux.
Although there is no equilibrium temperature gradient, significant fluxes
of both ion and electron heat are found.  Both heat fluxes have the same sign
as the particle flux, meaning heat flows in the $-\nabla n$ direction.

\begin{figure}[h!]
% Runs on edison in
% /global/scratch2/sd/landrema/20140730-01-coderunner-edison-subcritical-box
% Figure generated by first running
% m20140807_01_plotGS2_extractFluxesVsTimeForPaper.m on edison,
% then running
% m20140807_01_plotFluxesVsTimeForPaper.m on my laptop.
\includegraphics[width=3in]{m20140807_01_plotFluxesVsTimeForPaper.pdf}
%\includegraphics[width=3in]{fluxes.png}
\caption{(Color online)
Even in the absence of equilibrium temperature gradients, ion and electron heat fluxes
are obtained.
\label{fig:fluxes}}
\end{figure}

Using the amplitude of the normalized particle flux from figure \ref{fig:fluxes},
and converting to dimensional units, we obtain the following estimate for the turbulent diffusivity:
\begin{equation}
D \approx \left( 0.6 \mathrm{ \;m^2/s}\right)
\frac{\left(T / 500 \mathrm{\;eV}\right)^{3/2}}
{\left(B/2\mathrm{\;T}\right)^2 \left( L_n / 1 \mathrm{\;cm}\right)}
\end{equation}
This magnitude of diffusion coefficient is large enough to be of interest experimentally.

\section{Turbulence implies transient amplification of typical states}
\label{sec:proof}

It seems plausible that the previous two sections are related,
that a connection exists between transient linear amplification and sustained nonlinear turbulence.
Let us now prove this relationship rigorously.
Here we present two such arguments,
both of which yield the conclusion that non-modal amplification is a necessary condition
for sustained turbulence.  One argument for the case of hydrodynamics
was given in \cite{DelSoleNecessity} and section 4 of \cite{DelSoleSurvey},
and below we adapt this reasoning to gyrokinetics.  Before we do so, we give an alternative
argument that relies less on linear algebra.
For this discussion, it is useful to define a ``2D mode'' to be any structure with given $k_x$ and $k_y$,
but with no restriction on the $z$ dependence.

{\it Statement 1:} Consider a solution of the nonlinear system (\ref{eq:gke})-(\ref{eq:qn})
for which $W$ in (\ref{eq:W}) is not monotonically decreasing, and choose any
 time $t_0$ during which the free energy satisfies $dW/dt>0$.
(In sustained turbulence, where $W$ fluctuates about some mean value,
this condition must be satisfied some of the time.)  Suppose at $t_0$ we turn
off the nonlinear term
and evolve the system in time, as in figure \ref{fig:linearRestart}.
Then at least one 2D mode must grow linearly in the $W$ norm
for at least some time, i.e. there must be transient linear amplification.

{\it Proof:}
Recall that (\ref{eq:dWdt}) is a moment of both the nonlinear and linear dynamical equations.
When the nonlinear term is turned off at $t_0$, $S(\Phi,h_i,h_e)$ does not immediately change, because
the system state does not immediately change. Therefore, $dW/dt$ must vary \emph{continuously}
when the nonlinear term is turned off, and so $dW/dt$ is positive at $t_0$ under the linear dynamics.
Recall $W$ is a sum of positive contributions (\ref{eq:Wk}) from each 2D mode:
$W = \sum_{k_x,k_y} W_{k_x,k_y}$.
Therefore $dW_{k_x,k_y}/dt$ must be positive in the linear dynamics for at least one 2D mode,
so this mode must be growing transiently in the $W$ norm.

Roughly speaking, the upshot of this argument is that we should not be surprised there is transient amplification
after the nonlinear term is turned off in figure (\ref{fig:linearRestart}).
Such amplification \emph{must} occur.
Put another way, a necessary condition
for turbulence (in the absence of linear instability) is that typical states of the turbulence
can be transiently amplified. Furthermore, the energy growth in the 2D modes that are amplified
must be sufficient to balance the energy decay from the damped 2D modes.

Next, we present a more algebraic argument for the necessity of transient amplification, closely
following \cite{DelSoleNecessity}.  For this second approach, we let $\vect{h}$ denote
a column vector that specifies the state of the system $\{h_i,h_e\}$ in some convenient basis.
For example, $\vect{h}$ could represent the array of values specifying the distribution function in a code.

{\it Statement 2:} Given physical parameters for which
the system (\ref{eq:gke})-(\ref{eq:qn})
permits statistically steady turbulence,
there must exist a state $\vect{h}_0$ such that
(1) $dW/dt>0$ at $t=0$ when the system is evolved from $\vect{h}_0$, and
(2) the turbulent state $\vect{h}(t)$ has a nonzero average projection
onto $\vect{h}_0$.

{\it Proof:}
Let $\vect{L}$ denote the time-independent linear operator associated with (\ref{eq:gke})-(\ref{eq:qn})
in the absence of the nonlinear term,
so the linearized (\ref{eq:gke})-(\ref{eq:qn}) can be written
\begin{equation}
d\vect{h}/dt = \vect{L}\vect{h}.
\end{equation}
(Here we will take $\vect{L}$ to be real since the equations (\ref{eq:gke})-(\ref{eq:qn})
are real, though the following argument may be generalized to complex
$\vect{L}$ by replacing the transpose operation with Hermitian conjugation.)
Also let $\vect{E}$ be the time-independent operator that defines the free energy norm:
\begin{equation}
W = \vect{h}^{\transpose} \vect{E} \vect{h}
\label{eq:hEh}
\end{equation}
where $^\transpose$ denotes the transpose. Note that $\vect{E}$ is real and symmetric
($\vect{E} = \vect{E}^\transpose$). Differentiating (\ref{eq:hEh}),
\begin{equation}
\label{eq:dWdt2}
dW/dt = \vect{h}^{\transpose} \left[ (\vect{E}\vect{L})^\transpose + \vect{E}\vect{L} \right] \vect{h}
\end{equation}
which is equivalent to (\ref{eq:dWdt}). As discussed previously, the presence of the nonlinear term
does not alter (\ref{eq:dWdt}) (since the nonlinear term is annihilated in the integration over
the perpendicular directions $X_s$ and $Y_s$),
so (\ref{eq:dWdt2}) holds for both the linear and nonlinear gyrokinetic equations.
Introducing a time average $\left< \ldots \right>$ over statistically steady turbulence, then
\begin{equation}
\label{eq:avg}
\left< \vect{h}^{\transpose} \left[ (\vect{E}\vect{L})^\transpose + \vect{E}\vect{L} \right] \vect{h} \right>
=0.
\end{equation}

Next, observe that the operator $(\vect{E}\vect{L})^\transpose + \vect{E}\vect{L}$ in (\ref{eq:dWdt2})-(\ref{eq:avg})
is real and symmetric, so it possesses a complete set of orthogonal eigenvectors with real eigenvalues.
Let $\vect{u}_j$ and $\lambda_j$
denote these eigenvectors (normalized so $\vect{u}_j^\transpose \vect{u}_k = \delta_{j,k}$)
and their associated eigenvalues.
In the hydrodynamics literature, the eigenvectors $\vect{u}_k$ are sometimes called
instantaneous optimals; the $\vect{u}_j$ with maximum $\lambda_j$ is the fluctuation
that maximizes the instantaneous $dW/dt$.
We may decompose the state vector in the $\vect{u}_j$ basis:
$\vect{h}(t) = \sum_j a_j(t) \vect{u}_j$ for some amplitudes $a_j(t)$. Then
\begin{equation}
\label{eq:sum}
\sum_j \lambda_j \left< \left| a_j(t) \right|^2 \right>
=0.
\end{equation}
Neglecting the uninteresting case in which all terms in the sum (\ref{eq:sum}) are zero,
then there must be at least one positive eigenvalue $\lambda_{j_0}$ with an associated nonzero
average weight $\left< \left| a_{j_0}(t) \right|^2 \right>$. Considering the associated
$\vect{u}_{j_0}$ in (\ref{eq:dWdt2}), then this state $\vect{u}_{j_0}$ must give positive $dW/dt$,
proving statement 2.

Thus, we have showed in two ways that transient linear amplification is a necessary
condition for turbulence in the associated nonlinear system.  We have also shown that the amplitude of the
transiently growing state(s) must have a significant amplitude in the turbulent state, in the following senses.
In the first line of argument, not only must there be at least a single growing 2D mode,
but there must be enough growing 2D modes of sufficient amplitude to balance the negative $dW_{k_x,k_y}/dt$
of all the decaying 2D modes.
In the second argument, in (\ref{eq:sum})
there must not only be at least one state $j$ with an instantaneous growth rate
$\lambda_j >0$, but the amplitudes of states with positive and negative
$\lambda_j$ must be in balance so $\left< dW/dt \right>=0$. In a statistically steady state, there must be a balance between energy input from the instantaneously growing
linear states and the instantaneously decaying linear states.

Both of the arguments above relied on the fact that the norm $W$ satisfies the same equation
(\ref{eq:dWdt}) in both the linear and nonlinear dynamics.
Thus, the ``nonlinear invariant'' $W$ is a special norm, and the arguments do not generally apply to other norm-like quantities
such as $\int d^3r \left| \Phi \right|^2$ in which one might choose to measure transient amplification.
(In fact for any operator with damped eigenmodes,
even if transient amplification exists in one norm,
another norm always exists in which transient amplification is impossible.
The sum of squared eigenmode coefficients is such a norm \cite{FarrellIoannou1993}.)

While the arguments here relied on the gyrokinetic conservation law (\ref{eq:dWdt}), the same reasoning may be applied to
other models and systems in which an energy conservation equation holds. For example,
eq (6) in Ref. \cite{Drake} gives the appropriate conservation law for the fluid model
in that work, and the first line of (6) gives the associated norm to use in place of $W$.

\section{Discussion and conclusions}
\label{sec:conclusions}

The ``universal mode'' system (slab geometry with a density gradient, electrostatic,
no sheared equilibrium flow, and weak temperature gradients)
is known to have a dramatic change in linearly stability
as the magnetic shear is raised from zero to small finite values
\cite{Ross, Tsang, Antonsen,usUniversalInstability}.
As the eigenvalues are so sensitive to magnetic shear,
we might expect traditional modal analysis to
give a misleading picture of typical behavior,
as with other systems that have sensitive eigenvalues \cite{TrefethenEmbree}.
Indeed, here we have demonstrated that significant
transient linear amplification is possible in the drift wave system even when
all eigenmodes are decaying.
We have demonstrated that the amount and duration of amplification scale
with the magnetic shear length $L_s$, so these quantities naturally become
unbounded in the limit of vanishing shear.
The non-modal amplification also decreases with increasing collisionality,
particularly at the shortest perpendicular wavelengths,
and the amplification tends to be largest at wavelengths near $k_y \rho_i \sim 1-2$.
In contrast to more famous examples of non-modal amplification
from hydrodynamics \cite{TrefethenSubcritical}, and to some recent examples from plasma physics \cite{Newton, Alex, BarnesRotation, EdmundPRL},
the amplification in the plasma drift wave system here
does not depend on sheared equilibrium flow.

We have also demonstrated using nonlinear gyrokinetic simulations
that sustained turbulence is possible in this system
even when all linear modes are damped.
This phenomenon had been seen previously
in fluid models \cite{Scott1, Scott2, Drake}.
This phenomenon, clearly visible in figure \ref{fig:amplitudeDependence},
is opposite to the ``Dimits shift'' seen at different physical parameters,
in which nonlinear fluctuations
are suppressed despite the presence of linear instability.

In this system without unstable eigenvalues, we find that sustained nonlinear turbulence
and transient linear amplification both become possible around $L_s / L_n \gtrsim 20$
(for the collisionality $\nu_{ee} k_y \rho_i / \omega_* = 0.1$ and other parameters considered.)
As discussed in section \ref{sec:proof}, the fact that these two phenomena appear together is not
a coincidence. Transient linear amplification in the free energy norm $W$ is a necessary condition for statistically steady
turbulence, and the turbulent state must have a significant projection onto states that are linearly growing in the $W$
norm.  When the nonlinear term is turned off in a simulation of turbulence, as in figure \ref{fig:linearRestart},
amplification in $W$ norm must occur in at least one of the 2D modes.
The amplitude of the instantaneously growing states (instantaneous optimals) must be
significant, enough for the associated input of free energy to balance decay of energy
from the states that are decaying.

Note that the transient linear process emphasized here and nonlinear or secondary instability are not mutually
exclusive. Both are required to support turbulence.
Some form of secondary instability (as discussed in \cite{Drake}) must be present to transfer energy out of the transiently growing linear
structures before these structures begin to decay substantially. However, nonlinear processes cannot
directly contribute to the system's free energy $W$, due to the annihilation of the nonlinear term
in deriving (\ref{eq:dWdt}).

Several effects that may be important for realistic plasma turbulence
were not included in the present work, including toroidicity and trapped particles,
and electromagnetic fluctuations.
Therefore, further work is required to determine how relevant the nonlinear turbulence considered here
is for realistic experimental parameters.  However, we at least conclude that linear stability
need not preclude the possibility of significant turbulent particle and energy fluxes,
and that nonlinear simulations should sometimes be initialized at finite amplitude to examine this possibility.

\begin{acknowledgments}
This material is based upon work supported by the
U.S. Department of Energy, Office of Science, Office of Fusion Energy Science,
under Award Numbers DEFG0293ER54197 and DEFC0208ER54964.
Computations were performed on the Edison system at
the National Energy Research Scientific Computing Center, a DOE Office of Science User Facility supported by the Office of Science of the U.S. Department of Energy under Contract No. DE-AC02-05CH11231.
Conversations with 
Tom Antonsen, Michael Barnes, Greg Colyer, James Drake, Edmund Highcock, Greg Hammett, and Adil Hassam
are gratefully acknowledged.
\end{acknowledgments}

\appendix

\section{Linear stability in the absence of magnetic shear}
\label{appendix:zeroShear}

For reference, here we present the growth/damping rates and real frequencies of the
universal instability in the limit of vanishing
magnetic shear and vanishing collisions.
In the absence of shear, we can replace $\partial/\partial z$ in (\ref{eq:gke})
with $i k_{||}$ for some constant $k_{||}$.
Using (\ref{eq:qn}), and after some algebra, the local dispersion relation may then be written
\begin{eqnarray}
\frac{T_e}{T_i}-\zeta_i\frac{\omegase}{\omega}\left\{
-\left[\frac{\omega T_e}{\omegase T_i}+1\right]Z_i \Gamma_{0i}
+\eta_i Z_i \left[ \left(\frac{1}{2}+b_i\right) \Gamma_{0i}-b_i \Gamma_{1i}\right]
-\eta_i \zeta_i \left[1+\zeta_i Z_i\right] \Gamma_{0i}\right\}
&&\label{eq:dispersionRelation} \\
+
1+\zeta_e\frac{\omegase}{\omega}\left\{
\left[\frac{\omega}{\omegase}-1\right]Z_e \Gamma_{0e}
+\eta_e Z_e \left[ \left(\frac{1}{2}+b_e\right) \Gamma_{0e}-b_e \Gamma_{1e}\right]
-\eta_e \zeta_e \left[1+\zeta_e Z_e\right] \Gamma_{0e}\right\}
&=&0.
\nonumber
\end{eqnarray}
Here, $\zeta_s = \omega/(\left| k_{||} \right| v_s)$, $v_s = \sqrt{2 T_s/m_s}$ is the thermal speed, $Z_s = Z(\zeta_s)$, $Z$ is the plasma dispersion function,
$\Gamma_{js} = I_j(b_s)e^{-b_s}$, $I_j$ is a modified Bessel function, and $b_s = k_{\perp}^2 v_s^2/(2 \Omega_s^2)$.
As before, $\omega_* = k_y T_e/(e B L_n)$.
Figure \ref{fig:localDispersionRelation} shows the solution of (\ref{eq:dispersionRelation})
with largest imaginary part for the case $T_e = T_i$
and the deuterium-electron mass ratio.
We take $\eta_i = \eta_e = 0$
so it is certain that the ion-temperature-gradient (ITG) and electron-temperature-gradient (ETG)
modes are suppressed.
Figure \ref{fig:localDispersionRelation} was generated by applying nonlinear root-finding to (\ref{eq:dispersionRelation}),
verifying the solution agreed with initial-value gs2 simulations for several values of $k_{||}$ and $k_{\perp}$.
It can be seen that absolute linear instability exists for sufficiently small values of $\left|k_{||}\right|$
when $k_{\perp} \rho_i < 40$.

\begin{figure}[h!]
% Figure generated with m20141214_01_plotUniversalModeDispersionRelationForPaper.m
%\includegraphics[width=6in]{localLinearDispersionRelation.png}
\includegraphics[width=6in]{m20141214_01_plotUniversalModeDispersionRelationForPaper.pdf}
\caption{(Color online) Solution of the linear dispersion relation in the zero-shear (local) limit, i.e. the
root of (\ref{eq:dispersionRelation}) with largest imaginary part, for $\eta_i = \eta_e = 0$,
$T_e = T_i$, and $m_i = 3672 m_e$.
The (a) imaginary and (b)
real parts of the frequency are shown.
\label{fig:localDispersionRelation}}
\end{figure}

\bibliography{universalModeTurbulencePaper}

\end{document}

### 2014-12-16 universalModeTurbulencePaper.tex
Extended figure 5 to larger times
Implemented suggestions from Gabe

@@
 In contrast to hydrodynamic examples of non-modal growth and subcritical turbulence,
-here there is not sheared flow in the equilibrium.
+here there is no sheared flow in the equilibrium.
 Significant transient linear amplification
@@
 -\frac{f_{Ms}}{B L_n} \frac{\partial \left< \Phi\right>_{\vect{R}_s}}{\partial Y_s}
-\left[ 1 + \left( \energy-\frac{3}{2}\right) \eta_s\right]
+\left[ 1 + \left( \energy_s-\frac{3}{2}\right) \eta_s\right]
 \nonumber
@@
 Here $s \in \left\{i,e\right\}$ subscripts denote species,
-$\Phi(t,\vect{r})$ is the electrostatic potential, $h_s(t,\vect{R}_s,\energy,\mu)$ is the nonadiabatic distribution function,
-$\energy = v^2/2$, $\mu = v_{\bot}^2/(2B)$,
+$\Phi(t,\vect{r})$ is the electrostatic potential, $h_s(t,\vect{R}_s,\energy_s,\mu)$ is the nonadiabatic distribution function,
+$\energy_s = m_s v^2/(2 T_s)$, $\mu = v_{\bot}^2/(2B)$,
 and $f_{Ms}$ is the leading-order Maxwellian.
-The scale length of the equilibrium density $n$ is $L_n = -(1/n)(dn/dx)$,
+The scale length of the equilibrium density $n$ is $L_n = -n/(dn/dx)$,
 $q_i = e = -q_e$ is the proton charge,
@@
 Expanding the fluctuating quantities in spatial Fourier modes as $\Phi = \sum_{k_x,k_y} \Phi_{k_x,k_y}(t,z) \exp(i k_x x + i k_y y)$
-and $h_s = \sum_{k_x,k_y} h_{s,k_x,k_y}(t,z,\energy,\mu) \exp(i k_x X_s + i k_y Y_s)$,
+and $h_s = \sum_{k_x,k_y} h_{s,k_x,k_y}(t,z,\energy_s,\mu) \exp(i k_x X_s + i k_y Y_s)$,
 we may replace the gyro-averages with Bessel functions:
@@
 \begin{equation}
-\frac{d W}{d t} = S(\Phi,h_i,h_e)
+\frac{d W}{d t} = S(h_i,h_e)
 \label{eq:dWdt}
@@
 \label{eq:freeEnergySources}
-S(\Phi,h_i,h_e)
+S(h_i,h_e)
 &=&
@@
  \frac{T_s}{n_s}\frac{dn_s}{dx}
-\left[ 1 + \left( \energy-\frac{3}{2}\right) \eta_s\right]
+\left[ 1 + \left( \energy_s-\frac{3}{2}\right) \eta_s\right]
 \nonumber
@@
 As this instability is stabilized for $k_y \rho_i \ll 1$ when even a small amount of magnetic shear
-is introduced \cite{Ross, Tsang}, we might expect that this system can exhibit transient linear amplification, with an amount of
-amplification that tends to infinity as the magnetic shear tends to zero.  In this way, the apparently singular limit
+is introduced \cite{Ross, Tsang}, we might expect that this system can exhibit transient linear amplification, with
+the number of $e$-foldings of amplification tending to infinity as the magnetic shear tends to zero.  In this way, the apparently singular limit
 of zero shear may be seen as a continuous limit of the general shear case.
@@
 As transient amplification generally depends on the choice
-of norm, here we compare three reasonable (squared) norms: $W$, $\int dz |\Phi|^2$,
+of norm, here we compare three reasonable quadratic norms: $W$, $\int dz |\Phi|^2$,
 and the maximum of $|\Phi|^2$ along $z$.
@@
 and $\delta\! f_s =\sum_{k_x,k_y} \delta\! f_{s,k_x,k_y} \exp( i k_x x+i k_y y)$.
-Curves for the other $k_x$ in the simulation look similar to those in figure \ref{fig:linearBox}.
+Curves for the other $k_x$ in the simulation look similar to those in figure \ref{fig:linearBox},
+as expected from the $z$-translation symmetry of the system (\ref{eq:gke})-(\ref{eq:qn}); see discussion following (\ref{eq:kperp}).
 The amplitude of each
@@
 Linear modes with $k_y=0$ and $\partial/\partial z=0$ are damped only by collisions.
-(In the absence of any collisions, any $h_e$ and $h_i$ that depend only on $\energy$ and $\mu$ are steady-state solutions of the linearized system.)
+(In the absence of any collisions, any $h_e$ and $h_i$ that depend only on $\energy_s$ and $\mu$ are steady-state solutions of the linearized system.)
 Collisions introduce damping of these $k_y=0$ modes, with a damping rate that increases with $k_x^2$ due to the classical transport terms in the gyroaveraged collision operator.
@@
 Large-amplitude noise can be used, but this method is numerically inconvenient (at least in gs2) due to the rapid
-drop in amplitude over the first few timesteps due to strongly damped modes,
+drop in amplitude of strongly damped modes over the first few timesteps,
 which causes significant changes in the CFL-stable timestep associated
@@
 % Datasets are in /global/scratch2/sd/landrema/20141112-01-coderunner-edison-subcritical-box-lowLsOverLn
-% runs 54,55,58
+% runs 54,55,58,72
 % Figure is produced by first running
@@
 (a) When the initial amplitude is sufficient, turbulence grows and is sustained, whereas (b) for small initial amplitude,
-all modes decay as in a linear simulation.
+all modes decay as they do in a linear simulation.
 \label{fig:amplitudeDependence}}
@@
 
-{\it Statement 1:} Consider a solution of the nonlinear system (\ref{eq:gke})-(\ref{eq:qn})
-for which $W$ in (\ref{eq:W}) is not monotonically decreasing, and choose any
- time $t_0$ during which the free energy satisfies $dW/dt>0$.
-(In sustained turbulence, where $W$ fluctuates about some mean value,
-this condition must be satisfied some of the time.)  Suppose at $t_0$ we turn
-off the nonlinear term
-and evolve the system in time, as in figure \ref{fig:linearRestart}.
-Then at least one 2D mode must grow linearly in the $W$ norm
-for at least some time, i.e. there must be transient linear amplification.
+{\it Statement 1:} Consider a turbulent solution of the nonlinear gyrokinetic system (\ref{eq:gke})-(\ref{eq:qn}),
+i.e. a solution for which $W$ does not decay to zero at large $t$.
+Choose any time $t_0$ at which the free energy satisfies $dW/dt>0$.
+Suppose at $t_0$ we turn off the nonlinear term
+and re-evolve the system in time, as in figure \ref{fig:linearRestart}.
+Then at least one 2D mode must grow in the $W$ norm
+for some time following $t_0$, i.e. there must be transient linear amplification.
 
@@
 Recall that (\ref{eq:dWdt}) is a moment of both the nonlinear and linear dynamical equations.
-When the nonlinear term is turned off at $t_0$, $S(\Phi,h_i,h_e)$ does not immediately change, because
+When the nonlinear term is turned off at $t_0$, $S(h_i,h_e)$ does not immediately change, because
 the system state does not immediately change. Therefore, $dW/dt$ must vary \emph{continuously}
-when the nonlinear term is turned off, and so $dW/dt$ is positive at $t_0$ under the linear dynamics.
+when the nonlinear term is turned off, and so $dW/dt$ is positive for some period followiwng $t_0$ under the 
+action of the linear dynamics.
 Recall $W$ is a sum of positive contributions (\ref{eq:Wk}) from each 2D mode:
@@
 
-Roughly speaking, the upshot of this argument is that we should not be surprised there is transient amplification
+Roughly speaking, the upshot of this argument is that we should not be surprised that there is transient amplification
 after the nonlinear term is turned off in figure (\ref{fig:linearRestart}).
@@
 
-Next, we present a more algebraic argument for the necessity of transient amplification, closely
-following \cite{DelSoleNecessity}.  For this second approach, we let $\vect{h}$ denote
+Next, we present a second argument for the necessity of transient amplification.
+This argument is based on linear algebra and closely follows \cite{DelSoleNecessity}.  
+For this second approach, we let $\vect{h}$ denote
 a column vector that specifies the state of the system $\{h_i,h_e\}$ in some convenient basis.
@@
 
-{\it Statement 2:} Given physical parameters for which
-the system (\ref{eq:gke})-(\ref{eq:qn})
+{\it Statement 2:} Given a set of physical parameters for which
+the nonlinear gyrokinetic system (\ref{eq:gke})-(\ref{eq:qn})
 permits statistically steady turbulence,
-there must exist a state $\vect{h}_0$ such that
-(1) $dW/dt>0$ at $t=0$ when the system is evolved from $\vect{h}_0$, and
-(2) the turbulent state $\vect{h}(t)$ has a nonzero average projection
-onto $\vect{h}_0$.
+there exists at least one state $\vect{h}_0$ such that
+(1) $dW/dt>0$ if $\vect{h} = \vect{h}_0$, and
+(2) the squared projection of the turbulent state $\vect{h}(t)$ 
+onto $\vect{h}_0$, defined below, has a non-zero average.
 
@@
 We may decompose the state vector in the $\vect{u}_j$ basis:
-$\vect{h}(t) = \sum_j a_j(t) \vect{u}_j$ for some amplitudes $a_j(t)$. Then
+$\vect{h}(t) = \sum_j a_j(t) \vect{u}_j$ for some amplitudes $a_j(t)$. 
+Using this decomposition in (\ref{eq:avg}) gives
 \begin{equation}
@@
 This phenomenon, clearly visible in figure \ref{fig:amplitudeDependence},
-is opposite to the ``Dimits shift'' seen at different physical parameters,
+is opposite to the ``Dimits shift'' \cite{Dimits} seen at different physical parameters,
 in which nonlinear fluctuations

### 2014-12-20 universalModeTurbulencePaper.tex
Updated figure 7 and associated text; normalizing nu to v/L rather than omega*

@@
 no matter how small,
-the eigenmodes become linearly damped for $k_y \rho_i \ll 1$
+the eigenmodes become linearly damped for $k_y \rho_i \ll 1$,
+where $k_y$ is the perpendicular wavenumber and $\rho_i$ is the ion gyroradius
 \cite{Ross, Tsang, Antonsen}.
@@
 The system is uniquely specified by the following dimensionless parameters:
-$L_s / L_n$, $\eta_i$, $\eta_e$, $T_e / T_i$, and $k_y \rho_i \nu_{ee} / \omega_*$,
-where $\nu_{ee} = \sqrt{2}\pi n e^4 \ln\Lambda/[(4\pi\epsilon_0)^2 m_e^{1/2} T_e^{3/2}]$ is the electron-electron collision frequency,
-and $\omega_* = k_y T_e/(e B L_n)$ is the electron drift frequency.
+$L_s / L_n$, $\eta_i$, $\eta_e$, $T_e / T_i$, and collisionality $\nuee L_n / v_i$,
+where $\nuee = \sqrt{2}\pi n e^4 \ln\Lambda/[(4\pi\epsilon_0)^2 m_e^{1/2} T_e^{3/2}]$ is the electron-electron collision frequency
+and $v_i = \sqrt{2 T_i / m_i}$.
+%and $\omega_* = k_y T_e/(e B L_n)$ is the electron drift frequency.
 
@@
 a linear computation for a single $(k_x,k_y)$ in a case of very weak magnetic shear, $L_s / L_n = 300$.
-Other parameters used are $k_y \rho_i = 1$, $\nu_{ee} / \omega_* = 0.1$,
+Other parameters used are $k_y \rho_i = 1$ (where  $\rho_i = v_i / \Omega_i$), $\nuee L_n / v_i = 0.05$,
 $m_i = 3672 m_e$,
-$T_e = T_i$, and $\eta_i = \eta_e = 0$;
-the results are independent of $k_x$ as noted previously.
+$T_e = T_i$, and $\eta_i = \eta_e = 0$.
+The results are independent of $k_x$ up to a translation in $z$, as noted previously.
 Collisions are also included to eliminate absolute instability for $k_y \rho_i \ge 0.7$,
@@
 \caption{(Color online) (a) Linear evolution for $L_s/L_n=300$, $\eta_i = \eta_e = 0$,
-$\nu_{ee} = 0.1 \omega_*$, and $k_y \rho_i = 1$, illustrating transient amplification in three norms.
+$\nuee = 0.05 v_i/L_n$, and $k_y \rho_i = 1$, illustrating transient amplification in three norms.
 The initial conditions are detailed in the main text.
@@
 Results are shown for two values of collisionality,
-$\nuee k_y \rho_i / \omega_*= 0.2$ and 1.
+$\nuee L_n / v_i = 0.1$ and $0.5$.
 (When the collisionality is much smaller than these values,
@@
 consistent with the appearance of absolute instability in the limit of vanishing shear ($L_s \to \infty$).
-Parameters used are $k_y \rho_i = 2$ and $\nu_{ee} = 0.1 \omega_*/(k_y \rho_i)$.
+Parameters used are $k_y \rho_i = 2$ and $\nuee = 0.05 v_i / L_n$.
 \label{fig:transientAmplificationVsLsOverLn}}
@@
 for a range of $k_y$ surrounding $1/\rho_i$.  For this figure, $L_s/L_n=200$.
-The two collisionalities shown are $\nuee k_y \rho_i/\omega_* = 0.2$ and 1.
+The two collisionalities shown are $\nuee L_n / v_i = 0.1$ and $0.5$.
 \label{fig:transientAmplificationVsKy}}
@@
 For the demonstration here we consider
-the parameters $L_s/L_n = 20$, $T_e = T_i$, $m_i = 3672 m_e$, $\eta_i = \eta_e = 0$, and $\nu_{ee} k_y \rho_i / \omegase = 0.1$.
+the parameters $L_s/L_n = 20$, $T_e = T_i$, $m_i = 3672 m_e$, $\eta_i = \eta_e = 0$, and $\nuee L_n / v_i = 0.05$.
 We find this point in parameter-space is approximately on the edge of where sustained turbulence
@@
 to relax the equilibrium density gradient, and indeed we find such a flux.
-Although there is no equilibrium temperature gradient, significant fluxes
+Although there is no equilibrium temperature gradient, nonzero average fluxes
 of both ion and electron heat are found.  Both heat fluxes have the same sign
@@
 
+The fluxes in figure \ref{fig:fluxes} are normalized by gyro-Bohm values defined using the length $L_n$ and including the 2 in the thermal speed, e.g.
+the particle flux is normalized by $(\rho_i / L_n)^2 n v_i$.
+%Using the amplitude of the normalized particle flux from figure \ref{fig:fluxes},
+Converting to dimensional units for a deuterium plasma, we obtain the following estimate for the turbulent particle diffusivity:
+$D \approx \left( 0.4 \mathrm{ \;m^2/s}\right)
+\left(T / 500 \mathrm{\;eV}\right)^{3/2}
+\left(B/2\mathrm{\;T}\right)^{-2} \left( L_n / 1 \mathrm{\;cm}\right)^{-1}$.
+This magnitude of diffusion coefficient may be large enough to be of interest experimentally.
+A larger diffusion coefficient is obtained if $L_s/L_n$ is increased.
+
 \begin{figure}[h!]
+% Runs on edison in
+% /global/scratch2/sd/landrema/20141112-01-coderunner-edison-subcritical-box-lowLsOverLn
+% Figure generated by first running
+% m20141215_02_plotGS2_extractFluxesVsTimeForPaper.m on edison,
+% then running
+% m20141210_01_plotGs2FluxesVsTimeForPaper.m
+% on my laptop
+%------------------------------------------
+% Old version of figure:
 % Runs on edison in
@@
 % m20140807_01_plotFluxesVsTimeForPaper.m on my laptop.
-\includegraphics[width=3in]{m20140807_01_plotFluxesVsTimeForPaper.pdf}
+\includegraphics[width=3in]{m20141210_01_plotGs2FluxesVsTimeForPaper.pdf}
+%\includegraphics[width=3in]{m20140807_01_plotFluxesVsTimeForPaper.pdf}
 %\includegraphics[width=3in]{fluxes.png}
 \caption{(Color online)
-Even in the absence of equilibrium temperature gradients, ion and electron heat fluxes
+Fluxes for the simulation of Figure (\ref{fig:amplitudeDependence}).a with  $\nuee L_n / v_i = 0.05$ and
+$L_s/L_n=20$, approximately the maximum shear at which turbulence can be sustained at this collisionality.
+Even in the absence of equilibrium temperature gradients, nonzero average ion and electron heat fluxes
 are obtained.
@@
 \end{figure}
-
-Using the amplitude of the normalized particle flux from figure \ref{fig:fluxes},
-and converting to dimensional units, we obtain the following estimate for the turbulent diffusivity:
-\begin{equation}
-D \approx \left( 0.6 \mathrm{ \;m^2/s}\right)
-\frac{\left(T / 500 \mathrm{\;eV}\right)^{3/2}}
-{\left(B/2\mathrm{\;T}\right)^2 \left( L_n / 1 \mathrm{\;cm}\right)}
-\end{equation}
-This magnitude of diffusion coefficient is large enough to be of interest experimentally.
 
@@
 and transient linear amplification both become possible around $L_s / L_n \gtrsim 20$
-(for the collisionality $\nu_{ee} k_y \rho_i / \omega_* = 0.1$ and other parameters considered.)
+(for the collisionality $\nuee L_n / v_i = 0.05$ and other parameters considered.)
 As discussed in section \ref{sec:proof}, the fact that these two phenomena appear together is not
@@
 Here, $\zeta_s = \omega/(\left| k_{||} \right| v_s)$, $v_s = \sqrt{2 T_s/m_s}$ is the thermal speed, $Z_s = Z(\zeta_s)$, $Z$ is the plasma dispersion function,
-$\Gamma_{js} = I_j(b_s)e^{-b_s}$, $I_j$ is a modified Bessel function, and $b_s = k_{\perp}^2 v_s^2/(2 \Omega_s^2)$.
-As before, $\omega_* = k_y T_e/(e B L_n)$.
+$\Gamma_{js} = I_j(b_s)e^{-b_s}$, $I_j$ is a modified Bessel function, $b_s = k_{\perp}^2 v_s^2/(2 \Omega_s^2)$,
+and $\omega_* = k_y T_e/(e B L_n)$.
 Figure \ref{fig:localDispersionRelation} shows the solution of (\ref{eq:dispersionRelation})

### 2014-12-20 universalModeTurbulencePaper.tex
Tweak to abstract

@@
 and sustained nonlinear turbulence
-in a kinetic system that is linearly stable.
+in a kinetic system with no unstable eigenmodes.
 The particular system considered is an electrostatic slab
@@
 We prove these two phenomena are related:
-when sustained turbulence occurs without instability,
+when sustained turbulence occurs without unstable eigenmodes,
 states that are typical of the turbulence must

### 2014-12-27 universalModeTurbulencePaper.tex
Added figure 3b

@@
 % and then running
-% m20140728_01_plotTransientAmplificationKyScans.m on my laptop.
+% m20141122_01_plotTransientAmplificationKyScans.m on my laptop.
 % ----------------------------------------
@@
 \includegraphics[width=3in]{m20141122_01_plotTransientAmplificationKyScans.pdf}
+\includegraphics[width=3in]{m20141122_01_plotTransientAmplificationKyScans_times.pdf}
 %\includegraphics[width=3in]{m20140728_01_plotTransientAmplificationKyScans.pdf}
@@
 Conversations with 
-Tom Antonsen, Michael Barnes, Greg Colyer, James Drake, Edmund Highcock, Greg Hammett, and Adil Hassam
+Tom Antonsen, Michael Barnes, Greg Colyer, James Drake, Edmund Highcock, Greg Hammett, Adil Hassam, and Freja Nordsiek
 are gratefully acknowledged.

### 2015-01-04 universalModeTurbulencePaper.tex
Typo fix.

@@
 and
-$\left<\right>_{\vect{R}_s}$ and $\left<\right>_{\vect{r}}$ denote
-gyroaverages at fixed particle position $\vect{r}=(x,y,z)$ and guiding-center position $\vect{R}_s=(X_s,Y_s,Z_s)$.
+$\left<\right>_{\vect{r}}$ 
+and 
+$\left<\right>_{\vect{R}_s}$ 
+denote gyroaverages at fixed particle position $\vect{r}=(x,y,z)$ and guiding-center position $\vect{R}_s=(X_s,Y_s,Z_s)$.
 The distinction between $z$ and $Z_s$ may be neglected.

### 2015-01-09 universalModeTurbulencePaper.tex
Added new figure 2 relating finite-shear and zero-shear growth rates.

@@
 \newcommand{\sgn}{\mathrm{sgn}}
-\newcommand{\changed}[1]{#1}
-\newcommand{\todo}[1]{\textcolor{red}{#1}}
+\newcommand{\changed}[1]{\textcolor{red}{#1}}
 \newcommand{\transpose}{\mathrm{T}}
@@
 significant transient linear amplification is possible in this system,
-as long as the magnetic shear is not too strong.
+\changed{and that the transiently growing fluctuations
+resemble the zero-shear eigenmodes
+in their dependence of the growth rate on wavenumber.}
+%as long as the magnetic shear is not too strong.
 Second, in section \ref{sec:turbulence} we demonstrate that universal-mode turbulence without linear
@@
 in the absence of magnetic shear are given in Appendix \ref{appendix:zeroShear}.
+\changed{
+As detailed in the appendix, some insight into the instability can be obtained by expanding
+the the plasma dispersion function $Z$ for the electrons in the small-argument limit,
+$Z(\omega / (|k_{||}| v_e)) \approx i \sqrt{\pi}$ for $|\omega / (k_{||} v_e)| \ll 1$,
+and by expanding the ion $Z$ function in the large argument limit,
+$Z(\omega / (|k_{||}| v_i)) \approx - |k_{||}| v_i / \omega$ for $|\omega / (k_{||} v_i)| \gg 1$.
+One then finds a frequency with positive imaginary part $\propto i \sqrt{\pi} \omega_*$,
+indicating the instability is due to resonant electrons in the presence of a density gradient.}
 As this instability is stabilized for $k_y \rho_i \ll 1$ when even a small amount of magnetic shear
@@
 
+\changed{
+The transiently growing structures in the presence of magnetic shear
+are closely related to the zero-shear eigenmodes.  To demonstrate this relationship, we define an effective
+$k_{\perp}$ for the transient fluctuation at each time by
+\begin{equation}
+\label{eq:expectedKperp}
+\left< k_{\perp} \right>(t) = \frac{\int dz\; k_{\perp}(z) \left| \Phi(z,t)\right|^2}
+{\int dz\; k_{\perp}(z) \left| \Phi(z,t)\right|^2}.
+\end{equation}
+Similarly, we define an effective $|k_{||}|$ at each time for the transient structure
+by
+\begin{equation}
+\label{eq:expectedKpar}
+\left< \left| k_{||} \right| \right>(t) = \frac{\int dk_{||}  \left| k_{||}\right| \left| \tilde\Phi(k_{||},t)\right|^2}
+{\int dk_{||}\left| \tilde\Phi(k_{||},t)\right|^2}
+\end{equation}
+where $\tilde\Phi$ is the Fourier transform of $\Phi$ in $z$. Using these effective values
+of $k_{\perp}$ and $k_{||}$, the zero-shear dispersion relation may be evaluated.
+The resulting growth rate can then be compared to the instantaneous growth rate of the finite-shear
+transient structure, $\gamma = (2W(t))^{-1} dW/dt$.  This comparison is shown in figure \ref{fig:comparingTransientTo0Shear}.
+In this figure, we display 25 independent simulations of transients,
+varying $L_s/L_n$ between 20 and 300, and plotting the evolution of $\gamma$ from the time
+of peak growth rate to the time at which the energy begins to decay.
+In each case, there is very close agreement between the instantaneous growth rate of the transient
+and the `predicted' growth rate from the zero-shear dispersion relation.
+Thus, we can conclude that the transiently growing structures are closely related to the zero-shear
+eigenmodes.  Magnetic shear causes $k_{\perp}$ and $k_{||}$ to vary with time as the structure
+evolves (unlike the zero-shear case in which $k_{\perp}$ and $k_{||}$ are constants.)
+However, at each instant, the finite-shear transient grows at precisely the rate expected from the zero-shear
+instability at the average wavenumbers $\left< k_{\perp} \right>$ and $\left< \left| k_{||} \right| \right>$.
+For the simulations in figure  \ref{fig:comparingTransientTo0Shear}, $k_y \rho_i = 2$ and $\nuee = 0.05 v_i/L_n$.
+}
+
+\begin{figure}[h!]
+% Data generated on Edison in /global/scratch2/sd/landrema/20150109-01-coderunner-edison-0shearUniversalInstability/vnewk_e_0p05
+% Figure generated with m20150109_04_cmprngTransientGrowthRateTo0ShearForPaper.m
+\includegraphics[width=3in]{m20150109_04_cmprngTransientGrowthRateTo0ShearForPaper.pdf}
+\caption{(Color online)
+\changed{
+Although a change from zero to small finite magnetic shear causes a fundamental change in the eigenvalues,
+(which change from unstable to decaying), 
+the \emph{transient} finite-shear structures remain similar to the zero-shear modes.  For as shown here,
+the instantaneous growth rate of the transient finite-shear structure is nearly identical
+to the growth rate of a zero-shear eigenmode at the dominant $k_{\perp}$ and $k_{||}$ values
+(\ref{eq:expectedKperp})-(\ref{eq:expectedKpar}).  Each trajectory shown is an independent
+linear initial-value calculation, with $L_s/L_n$ varying from 20-300 between simulations.  The growth rate is plotted from
+the time of maximum growth rate (at the $+$ symbols) until decay begins.
+}
+\label{fig:comparingTransientTo0Shear}}
+\end{figure}
+
+\changed{
+For the range of $k_{||}$ and $k_{\perp}$ values computed for the transient structures
+of figure \ref{fig:comparingTransientTo0Shear}, the zero-shear dispersion relation
+is well approximated by the approximate analytic dispersion relation (\ref{eq:approxRoot}).  Thus, we can also conclude
+that the transient structures, like the zero-shear eigenmodes,
+are driven by the combination of resonant electrons and the density gradient.
+}
+
 Transient amplification is exponentially reduced by increasing magnetic shear,
@@
 % and then running
-% m20140723_03_plotTransientAmplificationLsOverLnScanForPaper.m on my laptop.
+% m20141121_03_plotTransientAmplificationLsOverLnScanForPaper.m on my laptop.
 % ----------------------------------------
@@
 all eigenmodes are decaying.
+\changed{
+As shown in figure \ref{fig:comparingTransientTo0Shear},
+the transient structures grow at the zero-shear
+growth rate, when evaluated at the dominant $k_{\perp}$ and $k_{||}$ at that instant.
+}
 We have demonstrated that the amount and duration of amplification scale
@@
 Conversations with 
-Tom Antonsen, Michael Barnes, Greg Colyer, James Drake, Edmund Highcock, Greg Hammett, Adil Hassam, and Freja Nordsiek
+Tom Antonsen, Michael Barnes, Greg Colyer, James Drake, Edmund Highcock, Greg Hammett, and Adil Hassam
 are gratefully acknowledged.
@@
 
+\changed{
+Some insight into the density-gradient-driven instability can be obtained by
+considering the limit $|\zeta_e| \ll 1$ so $Z_e \approx i\sqrt{\pi}$,
+and $|\zeta_i| \gg 1$ so $Z_i \approx -1/\zeta_i$. Considering solutions with $|\omega/\omega_*| \ll 1$,
+one finds
+\begin{equation}
+\label{eq:approxRoot}
+\omega = \frac{\omega_* \Gamma_{0i}}
+{\left(\frac{T_e}{T_i}+1\right)^2 + \left(\sqrt{pi}\frac{\omega_*}{|k_{||}| v_e}\right)^2}
+\left( \frac{T_e}{T_i} + 1 + i \sqrt{\pi}\frac{\omega_*}{|k_{||}| v_e}\right).
+\end{equation}
+The imaginary part of $\omega$ evidently arises due to the electron resonance $Z_e \approx i\sqrt{\pi}$.
+}
+
 \bibliography{universalModeTurbulencePaper}

### 2015-01-11 universalModeTurbulencePaper.tex
Typo fix
Added some new suggestions from Gabe.

@@
 resemble the zero-shear eigenmodes
-in their dependence of the growth rate on wavenumber.}
+in the dependence of their instantaneous growth rate on wavenumber.}
 %as long as the magnetic shear is not too strong.
@@
 -\frac{f_{Ms}}{B L_n} \frac{\partial \left< \Phi\right>_{\vect{R}_s}}{\partial Y_s}
-\left[ 1 + \left( \energy_s-\frac{3}{2}\right) \eta_s\right]
+\left[ 1 + \left( \energy_s-\frac{3}{2}\right) \eta_s\right],
 \nonumber
@@
 \begin{equation}
-\frac{d W}{d t} = S(h_i,h_e)
+\frac{d W}{d t} = S(h_i,h_e),
 \label{eq:dWdt}
@@
 indicating the instability is due to resonant electrons in the presence of a density gradient.}
-As this instability is stabilized for $k_y \rho_i \ll 1$ when even a small amount of magnetic shear
-is introduced \cite{Ross, Tsang}, we might expect that this system can exhibit transient linear amplification, with
-the number of $e$-foldings of amplification tending to infinity as the magnetic shear tends to zero.  In this way, the apparently singular limit
-of zero shear may be seen as a continuous limit of the general shear case.
+When an arbitrarily small amount of magnetic shear is introduced, the global eigenmodes are 
+no longer recognizable as the unsheared slab modes, and indeed are absolutely stable 
+for $k_y\rho_i \ll 1$ \cite{Ross,Tang}.  One might reasonably expect weakly sheared systems with 
+stable global eigenmodes to exhibit transient linear amplification, with the number of $e$-foldings 
+of amplification tending to infinity as the magnetic shear tends to zero.  In the following section, 
+we will show this hypothesis to be true, and thereby demonstrate the apparently singular limit of zero shear 
+to be a continuous limit of the general shear case.
+%As this instability is stabilized for $k_y \rho_i \ll 1$ when even a small amount of magnetic shear
+%is introduced \cite{Ross, Tsang}, we might expect that this system can exhibit transient linear amplification, with
+%the number of $e$-foldings of amplification tending to infinity as the magnetic shear tends to zero.  In this way, the apparently singular limit
+%of zero shear may be seen as a continuous limit of the general shear case.
 
@@
 The transiently growing structures in the presence of magnetic shear
-are closely related to the zero-shear eigenmodes.  To demonstrate this relationship, we define an effective
-$k_{\perp}$ for the transient fluctuation at each time by
+are closely related to the zero-shear eigenmodes.  To demonstrate this relationship, we consider a single
+$(k_x,k_y)$, and define an effective
+$k_{\perp}$ at each time for the transient fluctuation by
 \begin{equation}
 \label{eq:expectedKperp}
-\left< k_{\perp} \right>(t) = \frac{\int dz\; k_{\perp}(z) \left| \Phi(z,t)\right|^2}
-{\int dz\; k_{\perp}(z) \left| \Phi(z,t)\right|^2}.
+\left< k_{\perp} \right>(t) = \frac{\int dz\; k_{\perp}(z) \left| \Phi_{k_x,k_y}(t,z)\right|^2}
+{\int dz\; k_{\perp}(z) \left| \Phi_{k_x,k_y}(t,z)\right|^2}.
 \end{equation}
@@
 \label{eq:expectedKpar}
-\left< \left| k_{||} \right| \right>(t) = \frac{\int dk_{||}  \left| k_{||}\right| \left| \tilde\Phi(k_{||},t)\right|^2}
-{\int dk_{||}\left| \tilde\Phi(k_{||},t)\right|^2}
-\end{equation}
-where $\tilde\Phi$ is the Fourier transform of $\Phi$ in $z$. Using these effective values
+\left< \left| k_{||} \right| \right>(t) = \frac{\int dk_{||}  \left| k_{||}\right| \left| \tilde\Phi_{k_x,k_y}(t,k_{||})\right|^2}
+{\int dk_{||}\left| \tilde\Phi_{k_x,k_y}(t,k_{||})\right|^2},
+\end{equation}
+where $\tilde\Phi_{k_x,k_y}$ is the Fourier transform of $\Phi_{k_x,k_y}$ in $z$. Using these effective values
 of $k_{\perp}$ and $k_{||}$, the zero-shear dispersion relation may be evaluated.
@@
 \begin{equation}
-W = \vect{h}^{\transpose} \vect{E} \vect{h}
+W = \vect{h}^{\transpose} \vect{E} \vect{h},
 \label{eq:hEh}
@@
 \label{eq:dWdt2}
-dW/dt = \vect{h}^{\transpose} \left[ (\vect{E}\vect{L})^\transpose + \vect{E}\vect{L} \right] \vect{h}
+dW/dt = \vect{h}^{\transpose} \left[ (\vect{E}\vect{L})^\transpose + \vect{E}\vect{L} \right] \vect{h},
 \end{equation}
@@
 
-Note that the transient linear process emphasized here and nonlinear or secondary instability are not mutually
+Note that the transient linear process emphasized here and secondary instability are not mutually
 exclusive. Both are required to support turbulence.
@@
 \omega = \frac{\omega_* \Gamma_{0i}}
-{\left(\frac{T_e}{T_i}+1\right)^2 + \left(\sqrt{pi}\frac{\omega_*}{|k_{||}| v_e}\right)^2}
+{\left(\frac{T_e}{T_i}+1\right)^2 + \left(\sqrt{\pi}\frac{\omega_*}{|k_{||}| v_e}\right)^2}
 \left( \frac{T_e}{T_i} + 1 + i \sqrt{\pi}\frac{\omega_*}{|k_{||}| v_e}\right).

### 2015-01-13 universalModeTurbulencePaper.tex
Added suggestions from Gabe

@@
 \newcommand{\sgn}{\mathrm{sgn}}
-\newcommand{\changed}[1]{\textcolor{red}{#1}}
+%\newcommand{\changed}[1]{\textcolor{red}{#1}}
+\newcommand{\changed}[1]{#1}
 \newcommand{\transpose}{\mathrm{T}}
@@
 It is this latter situation that we will consider here.
-
-The fluid analogue of this drift wave system was investigated in \cite{Drake},
+(As an aside, the small-but-finite-shear eigenmodes
+are different from zero-shear modes not only for the
+universal instability, but also for the slab ion-temperature-gradient (ITG) instability.
+Specifically, in \cite{Plunk2014}, considering the slab mode by setting
+curvature drift effects $\omega_d$ to 0, the small-shear limit of (19) is different from 
+the zero shear result (12).)
+
+The fluid version of this drift wave system was investigated in \cite{Drake},
 using a model with three fluctuating quantities (density, electrostatic potential, and parallel velocity).
@@
 For the rest of this paper, we take $\eta_i = \eta_e = 0$
-so it is certain that the ion-temperature-gradient (ITG) and electron-temperature-gradient (ETG)
+so it is certain that the ion-temperature-gradient and electron-temperature-gradient (ETG)
 instabilities are suppressed.
@@
 no longer recognizable as the unsheared slab modes, and indeed are absolutely stable 
-for $k_y\rho_i \ll 1$ \cite{Ross,Tang}.  One might reasonably expect weakly sheared systems with 
+for $k_y\rho_i \ll 1$ \cite{Ross,Tsang}.  One might reasonably expect weakly sheared systems with 
 stable global eigenmodes to exhibit transient linear amplification, with the number of $e$-foldings 
@@
 from the states that are decaying.
+
+Considering the nonlinear and linear results together, the existence of transient amplification and 
+turbulence evidently depends on whether magnetic shear is sufficiently weak that un-sheared slab mode 
+dynamics can persist.  
+This geometric property can be controlled independently of the energy source of the turbulence (i.e. the density gradient),
+in contrast to turbulence driven by sheared flows where the nonuniformity of the flow plays two roles at once:
+shaping the global structure of the growing structures and providing energy to the turbulence.
+This feature makes the present system an interesting case study of sub-critical turbulence, and we are unaware of other such examples.”
 
@@
 Some insight into the density-gradient-driven instability can be obtained by
-considering the limit $|\zeta_e| \ll 1$ so $Z_e \approx i\sqrt{\pi}$,
+examining the limit $|\zeta_e| \ll 1$ so $Z_e \approx i\sqrt{\pi}$,
 and $|\zeta_i| \gg 1$ so $Z_i \approx -1/\zeta_i$. Considering solutions with $|\omega/\omega_*| \ll 1$,
@@
 \end{equation}
-The imaginary part of $\omega$ evidently arises due to the electron resonance $Z_e \approx i\sqrt{\pi}$.
+The imaginary part of $\omega$, proportional to  $Z_e \approx i\sqrt{\pi}$, evidently arises due to the electron resonance.
 }

### 2015-02-23 replyToReferee.txt
Implemented all suggestions of the referee except the first

We thank the referee for taking the time to carefully read the
manuscript and offer constructive feedback.  All of the
recommendations have been implemented, and can be identified in the
version of the new manuscript with changes in red. For each item in
the referee report, the relevant locations of the manuscript changes
are the following:

1) Page ??

2) Pages 8-9. Rather than add the suggested new figure as part (d) of
figure 1, we opted to add it as a new figure 2.

3) Page 16-17

i) Page 4

ii) Page 8

We hope that with these modifications, the manuscript is now ready for
publication.

### 2015-02-23 universalModeTurbulencePaper.tex
Implemented all suggestions of the referee except the first

@@
 \newcommand{\sgn}{\mathrm{sgn}}
-%\newcommand{\changed}[1]{\textcolor{red}{#1}}
-\newcommand{\changed}[1]{#1}
+\newcommand{\changed}[1]{\textcolor{red}{#1}}
+%\newcommand{\changed}[1]{#1}
 \newcommand{\transpose}{\mathrm{T}}
@@
 significant transient linear amplification is possible in this system,
-\changed{and that the transiently growing fluctuations
+and that the transiently growing fluctuations
 resemble the zero-shear eigenmodes
-in the dependence of their instantaneous growth rate on wavenumber.}
-%as long as the magnetic shear is not too strong.
+in the dependence of their instantaneous growth rate on wavenumber.
 Second, in section \ref{sec:turbulence} we demonstrate that universal-mode turbulence without linear
@@
 $T_s$ is the species temperature,
+\changed{$\eta_s=(d \ln T_s/dx)/(d \ln n/dx)$,}
 $C_s$ is the collision operator,
@@
 and $v_i = \sqrt{2 T_i / m_i}$.
-%and $\omega_* = k_y T_e/(e B L_n)$ is the electron drift frequency.
 
@@
 in the absence of magnetic shear are given in Appendix \ref{appendix:zeroShear}.
-\changed{
 As detailed in the appendix, some insight into the instability can be obtained by expanding
-the the plasma dispersion function $Z$ for the electrons in the small-argument limit,
+the plasma dispersion function $Z$ for the electrons in the small-argument limit,
 $Z(\omega / (|k_{||}| v_e)) \approx i \sqrt{\pi}$ for $|\omega / (k_{||} v_e)| \ll 1$,
@@
 $Z(\omega / (|k_{||}| v_i)) \approx - |k_{||}| v_i / \omega$ for $|\omega / (k_{||} v_i)| \gg 1$.
-One then finds a frequency with positive imaginary part $\propto i \sqrt{\pi} \omega_*$,
-indicating the instability is due to resonant electrons in the presence of a density gradient.}
+One then finds a frequency with positive imaginary part $\propto i \sqrt{\pi} / L_n$,
+indicating the instability is due to resonant electrons in the presence of a density gradient.
 When an arbitrarily small amount of magnetic shear is introduced, the global eigenmodes are
@@
 to be a continuous limit of the general shear case.
-%As this instability is stabilized for $k_y \rho_i \ll 1$ when even a small amount of magnetic shear
-%is introduced \cite{Ross, Tsang}, we might expect that this system can exhibit transient linear amplification, with
-%the number of $e$-foldings of amplification tending to infinity as the magnetic shear tends to zero.  In this way, the apparently singular limit
-%of zero shear may be seen as a continuous limit of the general shear case.
 
@@
 
-\changed{
 The transiently growing structures in the presence of magnetic shear
-are closely related to the zero-shear eigenmodes.  To demonstrate this relationship, we consider a single
+are related to the zero-shear eigenmodes.  To demonstrate this relationship, we consider a single
 $(k_x,k_y)$, and define an effective
 $k_{\perp}$ at each time for the transient fluctuation by
+\changed{
 \begin{equation}
 \label{eq:expectedKperp}
-\left< k_{\perp} \right>(t) = \frac{\int dz\; k_{\perp}(z) \left| \Phi_{k_x,k_y}(t,z)\right|^2}
-{\int dz\; k_{\perp}(z) \left| \Phi_{k_x,k_y}(t,z)\right|^2}.
-\end{equation}
+\left< k_{\perp} \right>(t) = 
+\frac{\int dz\; k_{\perp}(z) \left| \Phi_{k_x,k_y}(t,z)\right|^2}
+{\int dz\; \left| \Phi_{k_x,k_y}(t,z)\right|^2}.
+\end{equation}
+}
 Similarly, we define an effective $|k_{||}|$ at each time for the transient structure
@@
 of $k_{\perp}$ and $k_{||}$, the zero-shear dispersion relation may be evaluated.
-The resulting growth rate can then be compared to the instantaneous growth rate of the finite-shear
-transient structure, $\gamma = (2W(t))^{-1} dW/dt$.  This comparison is shown in figure \ref{fig:comparingTransientTo0Shear}.
+The resulting growth rate 
+\changed{
+and real frequency can then be compared to the behavior of the finite-shear
+transient structure.
+%$\gamma = (2W(t))^{-1} dW/dt$,
+An example of such a comparison is given in figure \ref{fig:comparingTransientTo0Shear_single},
+using the same simulation as in figure \ref{fig:transientAmplification}.
+For this case, we evaluate $\left< k_{\perp}\right>$ and $\left<|k_{||}|\right>$
+at $t = 300 L_n/v_i$.
+The figure shows that near this time, in the region $z \approx 300 L_n$ where the transient structure has greatest
+amplitude, the transient does resemble a wave packet with the real frequency and growth rate
+of the zero-shear eigenmode evaluated at $\left< k_{\perp}\right>$ and $\left<|k_{||}|\right>$.
+The zero-shear eigenmodes cannot remain true eigenmodes when even a minute amount of magnetic shear
+is introduced, since introducing shear turns $k_{\perp}$ from a constant into a function of $z$.
+But if this variation of $k_{\perp}(z)$ is on a longer scale length than $\sim 1/k_{||}$ of the perturbation, 
+then a localized wave packet can exist in a region of nearly constant $k_{\perp}$.
+We can then expect the zero-shear dispersion relation to give accurate predictions for the central frequency
+and growth rate of the packet. This packet will move in $z$, and as a result, the central values of $k_{\perp}$
+and $k_{||}$ will evolve, meaning that the equivalent zero-shear eigenmode at one time
+will differ from the equivalent zero-shear eigenmode at another time. 
+Since the frequency and growth rate of the finite-shear structure thus evolve in time, this structure is not
+itself an eigenmode, not until all points along the field line have had time to communicate with each other
+to set up a new global finite-shear eigenmode.
+However, at each instant of the transient growth phase, the finite-shear structure grows at precisely the rate expected from the zero-shear
+instability at the average wavenumbers $\left< k_{\perp} \right>$ and $\left< \left| k_{||} \right| \right>$.
+}
+
+\begin{figure}[h!]
+\includegraphics[width=4in]{m20150222_01_plotGS2TransientAmplificationVs0ShearEigenmode.pdf}
+\caption{(Color online)
+\changed{
+Comparison of the finite-shear transient from Figure \ref{fig:transientAmplification} to a
+zero-shear eigenmode, where $k_{\perp}$ and $k_{||}$ for the latter are chosen to be $\left<k_{\perp}\right>$ and $\left<|k_{||}|\right>$
+from the transient at $t=300 L_n/v_i$.
+It can be seen that near this time, the transient resembles a wave packet with roughly the same
+complex frequency as the zero-shear eigenmode.
+(The amplitude and phase of the eigenmode in (b) are arbitrary.)
+%from (\ref{eq:expectedKperp})-(\ref{eq:expectedKpar}).
+}
+\label{fig:comparingTransientTo0Shear_single}}
+\end{figure}
+
+\changed{
+To demonstrate this principle across a range of parameters,
+figure  \ref{fig:comparingTransientTo0Shear} demonstrates that the
+correspondence between the finite-shear transients and zero-shear eigenmodes
+is maintained throughout the evolution of the transient and for other
+values of magetic shear.}
+%This comparison is shown in figure \ref{fig:comparingTransientTo0Shear}.
 In this figure, we display 25 independent simulations of transients,
-varying $L_s/L_n$ between 20 and 300, and plotting the evolution of $\gamma$ from the time
+varying $L_s/L_n$ between 20 and 300, and plotting the evolution of $\gamma = (2W(t))^{-1} dW/dt$ from the time
 of peak growth rate to the time at which the energy begins to decay.
+For all simulations shown, $k_y \rho_i = 2$ and $\nuee = 0.05 v_i/L_n$.
 In each case, there is very close agreement between the instantaneous growth rate of the transient
@@
 Thus, we can conclude that the transiently growing structures are closely related to the zero-shear
-eigenmodes.  Magnetic shear causes $k_{\perp}$ and $k_{||}$ to vary with time as the structure
-evolves (unlike the zero-shear case in which $k_{\perp}$ and $k_{||}$ are constants.)
-However, at each instant, the finite-shear transient grows at precisely the rate expected from the zero-shear
-instability at the average wavenumbers $\left< k_{\perp} \right>$ and $\left< \left| k_{||} \right| \right>$.
-For the simulations in figure  \ref{fig:comparingTransientTo0Shear}, $k_y \rho_i = 2$ and $\nuee = 0.05 v_i/L_n$.
-}
+eigenmodes.  
+%Magnetic shear causes $k_{\perp}$ and $k_{||}$ to vary with time as the structure
+%evolves (unlike the zero-shear case in which $k_{\perp}$ and $k_{||}$ are constants.)
 
@@
 % Figure generated with m20150109_04_cmprngTransientGrowthRateTo0ShearForPaper.m
-\includegraphics[width=3in]{m20150109_04_cmprngTransientGrowthRateTo0ShearForPaper.pdf}
+\includegraphics[width=4in]{m20150109_04_cmprngTransientGrowthRateTo0ShearForPaper.pdf}
 \caption{(Color online)
-\changed{
 Although a change from zero to small finite magnetic shear causes a fundamental change in the eigenvalues,
@@
 the time of maximum growth rate (at the $+$ symbols) until decay begins.
-}
 \label{fig:comparingTransientTo0Shear}}
@@
 
-\changed{
 For the range of $k_{||}$ and $k_{\perp}$ values computed for the transient structures
@@
 are driven by the combination of resonant electrons and the density gradient.
-}
 
@@
 Large-amplitude noise can be used, but this method is numerically inconvenient (at least in gs2) due to the rapid
-drop in amplitude of strongly damped modes over the first few timesteps,
-which causes significant changes in the CFL-stable timestep associated
+drop in amplitude of strongly damped modes over the first few time steps,
+which causes significant changes in the CFL-stable time step associated
 with the perpendicular $\vect{E}\times\vect{B}$ speed,
@@
 is to initialize with one of the least-damped linear eigenmodes scaled up to appropriate amplitude, with noise in the other modes.
-We find this method to be particularly convenient since variation in the stable timestep
+We find this method to be particularly convenient since variation in the stable time step
 is minimized.
@@
 the particle flux is normalized by $(\rho_i / L_n)^2 n v_i$.
-%Using the amplitude of the normalized particle flux from figure \ref{fig:fluxes},
 Converting to dimensional units for a deuterium plasma, we obtain the following estimate for the turbulent particle diffusivity:
@@
 
+\changed{
+As mentioned earlier, for certain values of shear, collisionality, and initial perturbation amplitude,
+a third type of behavior was sometimes observed besides the two outcomes shown in figure \ref{fig:amplitudeDependence}:
+turbulence resembling figure  \ref{fig:amplitudeDependence}.a can persist for a finite time but eventually decay linearly
+as in figure  \ref{fig:amplitudeDependence}.b.
+This finite-lifetime plasma turbulence is reminiscent of turbulent puffs in neutral (Navier-Stokes) pipe flow.
+In neutral pipe flow it has been found \cite{Hof2006, Tel} that in fact
+the lifetime of turbulence puffs does not become infinite at a critical value of the relevant physical parameter (Reynolds number) but rather
+the lifetime varies with Reynolds number exponentially, so the turbulence lifetime can become longer than any reasonable observation
+time.
+An interesting open question (which we leave for future research) is whether a similar phenomenon may occur
+in the plasma system we consider here.
+}
+
 \section{Turbulence implies transient amplification of typical states}
@@
 the system state does not immediately change. Therefore, $dW/dt$ must vary \emph{continuously}
-when the nonlinear term is turned off, and so $dW/dt$ is positive for some period followiwng $t_0$ under the 
+when the nonlinear term is turned off, and so $dW/dt$ is positive for some period following $t_0$ under the
 action of the linear dynamics.
@@
 all eigenmodes are decaying.
-\changed{
 As shown in figure \ref{fig:comparingTransientTo0Shear},
@@
 growth rate, when evaluated at the dominant $k_{\perp}$ and $k_{||}$ at that instant.
-}
 We have demonstrated that the amount and duration of amplification scale
@@
 
-\changed{
 Some insight into the density-gradient-driven instability can be obtained by
@@
 The imaginary part of $\omega$, proportional to  $Z_e \approx i\sqrt{\pi}$, evidently arises due to the electron resonance.
-}
 

### 2015-02-24 universalModeTurbulencePaper.tex
Implemented some suggestions from Greg Hammett

@@
 and sustained nonlinear turbulence
-in a kinetic system with no unstable eigenmodes.
+in a kinetic plasma system with no unstable eigenmodes.
 The particular system considered is an electrostatic slab
@@
 it has been established that
-such analysis can in practice give misleading predictions
+eigenmode analysis can in practice give misleading predictions
 about the stability of various systems \cite{TrefethenEmbree}.
 In particular, turbulence may be sustained in a number of fluid systems
-despite the absence of any unstable eigenmodes.
-The first such systems known were certain sheared flow patterns in
+despite the absence of any unstable eigenmodes,
+a phenomenon that may be called ``subcritical turbulence''.
+The first systems known to display this behavior were certain sheared flow patterns in
 neutral fluids, such as plane Poiseuille or plane Couette flow
@@
 
-For the cases of turbulence without absolute linear stability in the presence of sheared flows,
+For the cases of turbulence without absolute linear instability in the presence of sheared flows,
 for both neutral fluids and plasmas, the phenomenon of transient linear amplification
@@
 This oscillation has been called the ``universal instability'' as it was at first
-mistakenly thought to be linearly unstable in any plasma with a density gradient.
+thought to be linearly unstable in any plasma with a density gradient.
 The oscillation is electrostatic in nature, and linear instability can indeed be found
-in slab geometry without magnetic shear or temperature gradients.
+in slab geometry without magnetic shear.
 It was later established that when any amount of magnetic shear is included in the model,
 no matter how small,
-the eigenmodes become linearly damped for $k_y \rho_i \ll 1$,
+the eigenmodes with $k_y \rho_i \ll 1$ become linearly damped,
 where $k_y$ is the perpendicular wavenumber and $\rho_i$ is the ion gyroradius
@@
 
-The fluid version of this drift wave system was investigated in \cite{Drake},
-using a model with three fluctuating quantities (density, electrostatic potential, and parallel velocity).
-In this model all linear modes were damped, yet nonlinearly, turbulence could be sustained.
-It was pointed out that an important secondary
+Fluid versions of this drift wave system were investigated in \cite{Scott1,Scott2,Drake}.
+%using a model with three fluctuating quantities (density, electrostatic potential, and parallel velocity).
+In these models, all linear modes were damped, yet nonlinearly, turbulence could be sustained.
+In \cite{Drake} it was pointed out that an important secondary
 instability in this system is a drift wave rotated by $\pi/2$ in the plane
@@
 include magnetic shear, which we will find to have a very strong effect on
-transient behavior.
+some aspects of the transient behavior.
 
@@
 First, in section \ref{sec:transient} we demonstrate that
-significant transient linear amplification is possible in this system,
-and that the transiently growing fluctuations
+significant transient linear amplification is possible in this system.
+The transiently growing fluctuations
 resemble the zero-shear eigenmodes
-in the dependence of their instantaneous growth rate on wavenumber.
+in the dependence of their instantaneous growth rate on wavenumber,
+though the transients are sensitive to magnetic shear
+in their global time-dependence.
 Second, in section \ref{sec:turbulence} we demonstrate that universal-mode turbulence without linear
-instability, seen previously in the fluid model in \cite{Drake},
+instability, seen previously in fluid models in \cite{Scott1,Scott2,Drake},
 is also possible in the gyrokinetic model.
@@
 %This comparison is shown in figure \ref{fig:comparingTransientTo0Shear}.
-In this figure, we display 25 independent simulations of transients,
+In this figure, we display 25 independent simulations of linear transients,
 varying $L_s/L_n$ between 20 and 300, and plotting the evolution of $\gamma = (2W(t))^{-1} dW/dt$ from the time
@@
 (which change from unstable to decaying),
-the \emph{transient} finite-shear structures remain similar to the zero-shear modes.  For as shown here,
+the \emph{transient} finite-shear structures retain a similarity to the zero-shear modes.  For as shown here,
 the instantaneous growth rate of the transient finite-shear structure is nearly identical
@@
 the major radius, $\hat{s} = (r/q) dq/dr$, and $r$ is the minor radius.)
-Figure \ref{fig:transientAmplificationVsLsOverLn}.b demonstrates that the time during which amplification occurs
+Figure \ref{fig:transientAmplificationVsLsOverLn}.b demonstrates that the time over which amplification occurs
 varies approximately linearly with $L_s/L_n$.
+The figure compares two measures of the duration of amplification, measuring to the time of maximum either from $t=0$
+or from the time of minimum $W$, akin to figure \ref{fig:transientAmplification}.b but in the time coordinate.
+These two measures of amplification duration display a similar linear trend.
+
+\changed{
+While figures \ref{fig:comparingTransientTo0Shear_single}-\ref{fig:comparingTransientTo0Shear}
+made the case that the transients were insensitive to shear in one sense,
+it is not a contradiction that the transients \emph{are} sensitive to shear
+in the sense of figures \ref{fig:transientAmplificationVsLsOverLn}.a-b.
+Figures  \ref{fig:comparingTransientTo0Shear_single}-\ref{fig:comparingTransientTo0Shear}
+show the dominant complex frequency of the transient at each instant depends on time
+only through $\left<k_{\perp}\right>(t)$
+and $\left<|k_{||}|\right>(t)$ and the zero-shear dispersion relation, with
+no significant additional dependence on shear.  However,  $\left<k_{\perp}\right>(t)$
+and $\left<|k_{||}|\right>(t)$ are sensitive to shear, yielding the sensitivity
+of the overall $W(t)$ curve for each transient seen in figures \ref{fig:transientAmplificationVsLsOverLn}.a-b.
+}
 
@@
 
-Throughout this section, the numerical parameters used were as follows: $12 \times 2$ grid points in pitch angle and $\sgn(v_{||})$, 12 grid points in energy, 24 grid points per cell in $z$ (with cells at $k_y=j \Delta k_y$ linked if separated in $k_x$ by $4 j \Delta k_x$), CFL number 0.5, $256 \times 64$ modes in $x$ and $y$ before dealiasing, and box size $(L_x, L_y, L_z) = (31.4 \rho_i, \; 62.8 \rho_i, \; 160 L_n)$.
-It was verified that both the linear and nonlinear results reported in this section were not significantly altered under factor-of-2 changes in any of these numerical parameters.
+Throughout this section, the numerical parameters used were as
+follows: $12 \times 2$ grid points in pitch angle and $\sgn(v_{||})$,
+12 grid points in energy, 24 grid points per cell in $z$ (with cells
+at $k_y=j \Delta k_y$ linked if separated in $k_x$ by $4 j \Delta k_x$
+for each integer $j\ge 1$), CFL number 0.5, $256 \times 64$ modes in $x$
+and $y$ before dealiasing, and box size $(L_x, L_y, L_z) = (31.4
+\rho_i, \; 62.8 \rho_i, \; 160 L_n)$.  It was verified that both the
+linear and nonlinear results reported in this section were not
+significantly altered under factor-of-2 changes in any of these
+numerical parameters.
 
@@
 % m20140807_01_plotFluxesVsTimeForPaper.m on my laptop.
-\includegraphics[width=3in]{m20141210_01_plotGs2FluxesVsTimeForPaper.pdf}
+\includegraphics[width=4in]{m20141210_01_plotGs2FluxesVsTimeForPaper.pdf}
 %\includegraphics[width=3in]{m20140807_01_plotFluxesVsTimeForPaper.pdf}
@@
 time.
-An interesting open question (which we leave for future research) is whether a similar phenomenon may occur
-in the plasma system we consider here.
+An interesting open question (which we leave for future research) is whether a similar scaling
+of the turbulence lifetime may be present in the plasma system we consider here.
 }
@@
 For this discussion, it is useful to define a ``2D mode'' to be any structure with given $k_x$ and $k_y$,
-but with no restriction on the $z$ dependence.
+but with no restriction on the $z$ dependence (so the structure need not be an eigenmode).
 
@@
 are real, though the following argument may be generalized to complex
-$\vect{L}$ by replacing the transpose operation with Hermitian conjugation.)
+$\vect{L}$ by replacing the transpose operation with Hermitian conjugation
+in the steps that follow.)
 Also let $\vect{E}$ be the time-independent operator that defines the free energy norm:
@@
 
-Thus, we have showed in two ways that transient linear amplification is a necessary
-condition for turbulence in the associated nonlinear system.  We have also shown that the amplitude of the
-transiently growing state(s) must have a significant amplitude in the turbulent state, in the following senses.
-In the first line of argument, not only must there be at least a single growing 2D mode,
-but there must be enough growing 2D modes of sufficient amplitude to balance the negative $dW_{k_x,k_y}/dt$
-of all the decaying 2D modes.
-In the second argument, in (\ref{eq:sum})
-there must not only be at least one state $j$ with an instantaneous growth rate
-$\lambda_j >0$, but the amplitudes of states with positive and negative
-$\lambda_j$ must be in balance so $\left< dW/dt \right>=0$. In a statistically steady state, there must be a balance between energy input from the instantaneously growing
-linear states and the instantaneously decaying linear states.
+Thus, we have showed in two ways that transient linear amplification
+is a necessary condition for turbulence in the associated nonlinear
+system.  We have also shown that the amplitude of the transiently
+growing state(s) must have a significant amplitude in the turbulent
+state, in the following senses.  In the first line of argument, not
+only must there be at least a single growing 2D mode, but there must
+be enough growing 2D modes of sufficient amplitude to balance the
+negative $dW_{k_x,k_y}/dt$ of all the decaying 2D modes.  In the
+second argument, in (\ref{eq:sum}) there must not only be at least one
+state $j$ with an instantaneous growth rate $\lambda_j >0$, but the
+amplitudes of states with positive and negative $\lambda_j$ must be in
+balance so $\left< dW/dt \right>=0$. In a statistically steady state,
+there must be a balance between energy input from the instantaneously
+growing linear states and energy extraction from the instantaneously
+decaying linear states.
 
@@
 The amplitude of the instantaneously growing states (instantaneous optimals) must be
-significant, enough for the associated input of free energy to balance decay of energy
-from the states that are decaying.
+significant, enough for the associated input of free energy to balance
+the states that are decaying.
 
@@
 shaping the global structure of the growing structures and providing energy to the turbulence.
-This feature makes the present system an interesting case study of sub-critical turbulence, and we are unaware of other such examples.”
+This feature makes the present system an interesting case study of subcritical turbulence, and we are unaware of other such examples.”
 
@@
 directly contribute to the system's free energy $W$, due to the annihilation of the nonlinear term
-in deriving (\ref{eq:dWdt}).
+in the derivation of (\ref{eq:dWdt}).
 
@@
 is for realistic experimental parameters.  However, we at least conclude that linear stability
-need not preclude the possibility of significant turbulent particle and energy fluxes,
-and that nonlinear simulations should sometimes be initialized at finite amplitude to examine this possibility.
+need not preclude the possibility of significant turbulent particle and energy fluxes.
+%and that nonlinear simulations should sometimes be initialized at finite amplitude to examine this possibility.
 
@@
 This material is based upon work supported by the
-U.S. Department of Energy, Office of Science, Office of Fusion Energy Science,
-under Award Numbers DEFG0293ER54197 and DEFC0208ER54964.
+U.S. Department of Energy, Office of Fusion Energy Science,
+under Award Numbers DEFG0293ER54197 and DEFC0208ER54964,
+and the SciDAC Center for the Study of Plasma Microturbulence.
 Computations were performed on the Edison system at
@@
 
-\section{Linear stability in the absence of magnetic shear}
+\section{Linear instability in the absence of magnetic shear}
 \label{appendix:zeroShear}

### 2015-02-26 replyToReferee.txt
Now responding to referee's comment 1

@@
 We thank the referee for taking the time to carefully read the
-manuscript and offer constructive feedback.  All of the
-recommendations have been implemented, and can be identified in the
-version of the new manuscript with changes in red. For each item in
-the referee report, the relevant locations of the manuscript changes
-are the following:
+manuscript and offer constructive feedback.
 
-1) Page ??
+Regarding suggestion #1, discussing the formal mathematical
+relationship between sensitivity of eigenmodes and transient growth
+using concepts of pseudospectra:
+
+We have added a paragraph discussing this subject to the end of
+Section III.  However, we are reluctant to assert too strong a
+connection to the mathematical theory of pseudospectra due to a
+technical complication.  In the 'destabilizing perturbations' section
+of Trefethen to which you refer, eq (13), the linear operator is given
+a perturbation of norm bounded by some finite epsilon, causing the
+eigenvalues to move by an amount that continuously goes to 0 as
+epsilon -> 0. However, all evidence suggests that when shear is
+changed from 0 to any finite value in the universal instability,
+eigenvalues change discontinuously, with the change not going to 0
+continuously as Ls/Ln -> infinity. In particular, the eigenvalues
+discontinuously change from unstable to stable for a range of k_y, as
+discussed in Refs. [25]-[26]. The reason for this discontinuity is
+presumably the fact that the z domain must be infinite to resolve
+eigenmode structures as shear goes to zero.  For any nonzero shear,
+there is always some z for which the argument of J_0 is changed
+significantly compared to the zero-shear case. Thus, to analyze the
+role of magnetic shear in the universal instability using the formal
+theory of pseudospectra, a level of mathematical care is required that
+is beyond the scope of the present manuscript.
+
+All of the other recommendations of the referee have been implemented,
+and can be identified in the version of the new manuscript with
+changes in red. For each item in the referee report, the relevant
+locations of the manuscript changes are the following:
 

### 2015-02-26 universalModeTurbulencePaper.tex
Now responding to referee's comment 1

@@
 \author{Gabriel G. Plunk}
-\affiliation{Max Planck Institute for Plasma Physics, EURATOM Association and Max-Planck/Princeton Research Center for Plasma Physics}
+\affiliation{Max Planck Institute for Plasma Physics, Max-Planck/Princeton Research Center for Plasma Physics}
 \author{William Dorland}
@@
 correspondence between the finite-shear transients and zero-shear eigenmodes
-is maintained throughout the evolution of the transient and for other
+is maintained throughout the evolution of the transient, for a range of
 values of magetic shear.}
@@
 \end{figure}
+
+\changed{In the context of sheared flows in neutral fluids, a
+  relationship between transient growth and eigenvalue sensitivity has
+  been formally proved using the notion of a pseudospectrum
+  \cite{ReddyHenningson,TrefethenSubcritical,TrefethenEmbree}. The
+  $\epsilon$-pseudospectrum of a linear operator can be defined as the
+  union of the sets of eigenvalues obtained when the operator is
+  perturbed by any other operator of norm $\le \epsilon$, providing a
+  precise notion of the sensitivity of eigenvalues.  One can
+  rigorously bound the amount of possible transient growth by the
+  extent of the pseudospectrum into the unstable half of the complex
+  plane \cite{ReddyHenningson} (eq (12) of
+  \cite{TrefethenSubcritical}.)  However, for the universal
+  instability, it is not clear that changing the magnetic shear from 0
+  to a small finite value has the form of the operator perturbation
+  used to define the pseudospectrum.  The reason is that the $z$
+  domain is infinite, and for any nonzero shear, there is always some
+  $z$ for which the argument of $J_0$ is changed significantly
+  compared to the zero-shear case.  Indeed, when the shear changes
+  from 0 to any small finite value, the eigenvalues (for a range of
+  $k_y$) appear to change from unstable to stable discontinuously
+  rather than continuously \cite{Ross,Tsang}.  Thus, it may or may not
+  be possible to analyze the role of magnetic shear in the universal
+  instability using the formal theory of pseudospectra, and we leave
+  this careful mathematical analysis for future work.  }
 

### 2015-03-20 universalModeTurbulencePaper.tex
Updates for referee response

@@
 \newcommand{\sgn}{\mathrm{sgn}}
-\newcommand{\changed}[1]{\textcolor{red}{#1}}
-%\newcommand{\changed}[1]{#1}
+%\newcommand{\changed}[1]{\textcolor{red}{#1}}
+\newcommand{\changed}[1]{#1}
 \newcommand{\transpose}{\mathrm{T}}
@@
 The distinction between $z$ and $Z_s$ may be neglected.
+\changed{
+We consider a flux tube domain, in which the fluctuating quantities can be taken
+to be periodic in $x$ and $y$.
+}
 Expanding the fluctuating quantities in spatial Fourier modes as $\Phi = \sum_{k_x,k_y} \Phi_{k_x,k_y}(t,z) \exp(i k_x x + i k_y y)$
@@
 = \int d^3 r \int d^3v \left< \ldots\right>_{\vect{r}}
-$ are used.)
+$ are used.
+\changed{
+The $v_{||} \partial h_s/\partial z$ term in (\ref{eq:gke}) has been integrated by parts;
+as detailed on page 66 of \cite{BanonNavarroThesis}, the associated boundary term vanishes for a finite $z$ domain
+using a twist-and-shift parallel boundary condition \cite{Beer1995},
+taking the fluctuation amplitude to vanish at the ends of each set of linked domains.)}
 The first term on the right-hand side of (\ref{eq:freeEnergySources})
@@
 
-Results for the three norms shown are quite similar,
-though the curve for the $W$ norm appears slightly higher than the other curves
-due to less decay in the initial period $t v_i / L_n <20$.
-Indeed, we find
-all results described or plotted for the remainder of this paper are quite similar
-for all of these norms.
-Henceforth we restrict our attention to the $W$ norm since it is well motivated
-by the conservation equation (\ref{eq:dWdt}).
-
-For the calculation in figure \ref{fig:transientAmplification}, the initial condition used
-is the standard ``noise'' initial condition of gs2:
-$g_s = h_s - \left< \Phi\right>_{\vect{R}_s} q_s f_{Ms} / T_s$ is set to a Maxwellian velocity distribution
-times a random complex number at each grid point in $z$.
-(The real and imaginary parts of these complex numbers are each taken from a uniform distribution on $[-1,1]$.)
-While techniques exist to calculate the optimal perturbation
-that gives maximum amplification \cite{FarrellIoannou, SquirePRL, SquireApJ}, such techniques
-are numerically challenging for this kinetic model due to the high dimensionality
-of the problem.
-The noise initial condition is a convenient shortcut, ensuring energy is likely
-to be given to a growing perturbation if one exists.
-It has been shown in \cite{Camargo, SquireApJ} that for studies of transient linear
-amplification, random initial conditions are a reasonable proxy
+For the calculation in figure \ref{fig:transientAmplification}, the
+initial condition used is the standard ``noise'' initial condition of
+gs2: $g_s = h_s - \left< \Phi\right>_{\vect{R}_s} q_s f_{Ms} / T_s$ is
+set to a Maxwellian velocity distribution times a random complex
+number at each grid point in $z$.  (The real and imaginary parts of
+these complex numbers are each taken from a uniform distribution on
+$[-1,1]$.)  While techniques exist to calculate the optimal
+perturbation that gives maximum amplification \cite{FarrellIoannou,
+  SquirePRL, SquireApJ}, such techniques are numerically challenging
+for this kinetic model due to the high dimensionality of the problem.
+The noise initial condition is a convenient shortcut, ensuring energy
+is likely to be given to a growing perturbation if one exists.  It has
+been shown in \cite{Camargo, SquireApJ} that for studies of transient
+linear amplification, random initial conditions are a reasonable proxy
 for a directly calculated optimal initial condition.
+\changed{Figure \ref{fig:transientAmplification}.b shows that 
+when the calculation is repeated many times, the dynamics of the energy in figure \ref{fig:transientAmplification}.a are typical.}
+
+Results for the three norms shown in figure
+\ref{fig:transientAmplification}.a are quite similar, though the curve
+for the $W$ norm appears slightly higher than the other curves due to
+less decay in the initial period $t v_i / L_n <20$.  Indeed, we find
+all results described or plotted for the remainder of this paper are
+quite similar for all of these norms, 
+\changed{using the locally-Maxwellian
+initial perturbations described above.  However, there are certain
+non-Maxwellian initial perturbations that can cause arbitrarily fast
+growth in the electric field energy norms but not in the $W$ norm \cite{GregIan}.
+Furthermore, it is $W$ that is conserved by the nonlinearities (recall
+the conservation equation (\ref{eq:dWdt})) and thus the $W$ norm that
+plays a role in determining bounds on the possibility of
+nonlinearly-sustained turbulence, as we will show in section
+\ref{sec:proof}.  For these reasons, we henceforth restrict our attention
+to the $W$ norm.}
 
@@
 \includegraphics[width=3in]{m20141119_01_plotGS2TransientAmplificationForPaper_figureA_a.pdf}
+\includegraphics[width=3in]{m20150305_02_plotManyIndntclGS2TransntAmplifctnRuns.pdf}
 % Figure generated with m20141121_01_plotGS2TransientAmplificationForPaper.m
-\includegraphics[width=3in]{m20141121_01_plotGS2TransientAmplificationForPaper_figuresBC_b.pdf}
 \includegraphics[width=3in]{m20141121_01_plotGS2TransientAmplificationForPaper_figuresBC_c.pdf}
+\includegraphics[width=3in]{m20141121_01_plotGS2TransientAmplificationForPaper_figuresBC_d.pdf}
 \caption{(Color online) (a) Linear evolution for $L_s/L_n=300$, $\eta_i = \eta_e = 0$,
@@
 The initial conditions are detailed in the main text.
-(b) The amount of transient amplification may be defined relative to $t=0$ or relative
+\changed{(b) Repeating the calculation 50 times, the results are always qualitatively similar.}
+(c) The amount of transient amplification may be defined relative to $t=0$ or relative
 to the minimum that occurs shortly thereafter.
-(c) The potential structure along a field line for this example.
+(d) The potential structure along a field line for this example.
 \label{fig:transientAmplification}}
@@
 
-Figure \ref{fig:transientAmplification}.b shows the same computation, zooming in to the initial transient period.
+Figure \ref{fig:transientAmplification}.c shows the same computation, zooming in to the initial transient period.
 Before the amplification begins around $t v_i/L_n=20$, there is an initial decay period due to strongly
@@
 For all simulations shown, $k_y \rho_i = 2$ and $\nuee = 0.05 v_i/L_n$.
-In each case, there is very close agreement between the instantaneous growth rate of the transient
+\changed{
+Both $\left<|k_{||}|\right>$ and $\left< k_\perp \right>$ increase with time during the transient
+growth, corresponding to movement upward and rightward in figure \ref{fig:localDispersionRelation}.a.
+As a result, the effective zero-shear growth rate decreases over time.
+}
+For all transients we examined, there is very close agreement between the instantaneous growth rate of the transient
 and the `predicted' growth rate from the zero-shear dispersion relation.
@@
 \begin{figure}[h!]
-% Data generated on Edison in /global/scratch2/sd/landrema/20150109-01-coderunner-edison-0shearUniversalInstability/vnewk_e_0p05
+% Data for finite-shear transients were generated on Edison in
+% /global/scratch2/sd/landrema/20141121-01-coderunner-edison-transientAmplificationScans/aky2/
+%
+% Data for zero-shear dispersion relation were generated on Edison in 
+% /global/scratch2/sd/landrema/20150109-01-coderunner-edison-0shearUniversalInstability/vnewk_e_0p05
+%
 % Figure generated with m20150109_04_cmprngTransientGrowthRateTo0ShearForPaper.m
@@
 are driven by the combination of resonant electrons and the density gradient.
+
+\changed{ In contrast to the agreement shown in figure
+  \ref{fig:comparingTransientTo0Shear} for the amplification phase
+  following the maximum growth rate, we do not find such close
+  agreement between the zero-shear dispersion relation and the
+  finite-shear dynamics for very early times, before the maximum
+  growth rate occurs.  In this initial period, a single dominant wave
+  packet has not yet overtaken other fluctuations present in the
+  initial random noise.  We also do not see such close agreement
+  between the zero-shear and finite-shear behavior during the decay
+  phase of the latter, i.e. at large times.  The finite-shear decay
+  rate as $t \to \infty$ corresponds to an eigenvalue, and it was
+  established in previous work such as \cite{Ross,Tsang,Antonsen} that
+  the finite-shear eigenvalues are not continuously related to the
+  zero-shear eigenvalues.  In an eigenmode, points at all $z$ have a
+  coherent time-dependence.  Thus, this damping rate is determined by
+  a coherent interaction between points of different
+  $k_\perp(z)=k_y\sqrt{1+(z/L_s)^2}$, and between all $k_{||}$ Fourier
+  components.  This damping rate is thus not well approximated by the
+  zero-shear dispersion relation which assumes a single $k_\perp$ and
+  single $k_{||}$. }
 
@@
 for the hydrodynamic case.)
-However, the kinetic problem here is rather different
-from the hydrodynamic case in the sensitivity of the transient amplification
-to the geometric parameter $L_s/L_n$,
-which is not associated with entropy-producing collisions.
 
@@
   $z$ for which the argument of $J_0$ is changed significantly
-  compared to the zero-shear case.  Indeed, when the shear changes
-  from 0 to any small finite value, the eigenvalues (for a range of
-  $k_y$) appear to change from unstable to stable discontinuously
-  rather than continuously \cite{Ross,Tsang}.  Thus, it may or may not
-  be possible to analyze the role of magnetic shear in the universal
-  instability using the formal theory of pseudospectra, and we leave
-  this careful mathematical analysis for future work.  }
+  compared to the zero-shear case.  Indeed, all evidence suggests that
+  when the shear changes from 0 to any small finite value, the
+  eigenvalues (for a range of ky) change from unstable to stable
+  discontinuously rather than continuously \cite{Ross,Tsang}.  Thus,
+  it may or may not be possible to understand the role of magnetic shear
+  in the universal instability using the formal theory of
+  pseudospectra, and we leave this careful mathematical analysis for
+  future work.  }
 
@@
 is minimized.
+\changed{
+(In both simulations shown in figure  \ref{fig:amplitudeDependence}, the $k_y \rho_i=0$ Fourier component is initialized to 0,
+but this component (and the $k_y \rho_i=\pm 0.2$ components) grow up quickly for the following reason. Since observable quantities are real, there is a large $k_y \rho_i=-0.1$
+Fourier component present in the initial condition which is the complex conjugate of the large $k_y \rho_i=+0.1$ component shown,
+and these two components interact nonlinearly to drive components at the sum and difference wave numbers.)
+}
 
@@
 must be sufficient to balance the energy decay from the damped 2D modes.
+
+\changed{ We note that while $dW/dt$ is continuous when the nonlinear term
+  is turned off, each $dW_{k_x,k_y}/dt$ is generally not. The reason is that in order to annihilate the
+  nonlinear term in the derivation of (\ref{eq:dWdt})-(\ref{eq:freeEnergySources}), it is necessary to
+  integrate over $x$ and $y$, i.e. sum over $k_x$ and $k_y$. Thus, while (\ref{eq:dWdt}) is
+  identical for the nonlinear and linear dynamics, the analogous
+  equation for $dW_{k_x,k_y}/dt$ is different for the nonlinear and linear
+  dynamics. Removing the nonlinear term therefore causes a
+  discontinuous change in $dW_{k_x,k_y}/dt$.
+}
 

### 2015-04-09 universalModeTurbulencePaper.tex
Formatted for JPP

--- 2015-02-24 universalModeTurbulencePaper.tex
@@
-%\documentclass{amsart}
-%\documentclass[aip,graphicx]{revtex4-1}
-\documentclass[12pt,superscriptaddress]{revtex4}
-\usepackage{latexsym, color, graphicx}
-%\usepackage{latexsym, color, graphicx, comment}
-\usepackage{amssymb}
-\usepackage{amsmath}
-
-%\graphicspath{{../MATLAB/}}
-
-\newcommand{\red}[1]{\textcolor{red}{#1}}
-\newcommand{\vect}[1]{\mbox{\boldmath $#1$}}
-\newcommand{\erf}{\mathrm{erf}}
-\newcommand{\PS}{Pfirsch-Schl\"{u}ter~}
-\newcommand{\ExB}{\vect{E}\times\vect{B}}
-\newcommand{\Cnl}{C_{\mathrm{nl}a}}
-\newcommand{\vma}{\vect{v}_{\mathrm{m}a}}
-\newcommand{\vm}{\vect{v}_{\mathrm{m}}}
-\newcommand{\vda}{\vect{v}_{\mathrm{d}a}}
-\newcommand{\vE}{\vect{v}_{E}}
-\newcommand{\vdao}{\vect{v}_{\mathrm{d}a0}}
-\newcommand{\vdo}{\vect{v}_{\mathrm{d}0}}
-\newcommand{\vEo}{\vect{v}_{E0}}
-\newcommand{\vpar}{v_{||}}
-\newcommand{\vperp}{v_{\bot}}
-\newcommand{\rhop}{\rho_{\theta}}
-\newcommand{\rT}{r_{T}}
-\newcommand{\Ti}{T_{\mathrm{i}}}
-\newcommand{\rTi}{r_{T\mathrm{i}}}
-\newcommand{\rTe}{r_{T\mathrm{e}}}
-\newcommand{\reta}{r_{\eta}}
-\newcommand{\rn}{r_{n}}
-\newcommand{\rperp}{r_\bot}
-\newcommand{\fM}{f_{\mathrm{M}}}
-\newcommand{\fMa}{f_{\mathrm{M}a}}
-\newcommand{\fMi}{f_{\mathrm{M}i}}
-\newcommand{\fMe}{f_{\mathrm{Me}}}
-\newcommand{\fe}{f_{\mathrm{e}}}
-\newcommand{\Te}{T_{\mathrm{e}}}
-\newcommand{\vth}{v_{\mathrm{th}}}
-\newcommand{\nuii}{\nu_{ii}}
-\newcommand{\nuPrime}{\hat{\nu}}
-\newcommand{\gradpar}{\nabla_{||}}
-\newcommand{\kpar}{k_{||}}
-\newcommand{\psia}{\psi_0}
-\newcommand{\psiaHat}{\hat{\psi}_0}
+% Copyright (C) 2011 Cambridge University Press
+
+\NeedsTeXFormat{LaTeX2e}
+
+\documentclass{jpp}
+
+\usepackage{graphicx}
+\usepackage{natbib}
+
+% See if the author has AMS Euler fonts installed: If they have, attempt
+% to use the 'upmath' package to provide upright math.
+\ifCUPmtlplainloaded \else
+  \checkfont{eurm10}
+  \iffontfound
+    \IfFileExists{upmath.sty}
+      {\typeout{^^JFound AMS Euler Roman fonts on the system,
+                   using the 'upmath' package.^^J}%
+       \usepackage{upmath}}
+      {\typeout{^^JFound AMS Euler Roman fonts on the system, but you
+                   dont seem to have the}%
+       \typeout{'upmath' package installed. JPP.cls can take advantage
+                 of these fonts, if you use 'upmath' package.^^J}%
+       \providecommand\upi{\pi}%
+      }
+  \else
+    \providecommand\upi{\pi}%
+  \fi
+\fi
+
+% See if the author has AMS symbol fonts installed: If they have, attempt
+% to use the 'amssymb' package to provide the AMS symbol characters.
+
+\ifCUPmtlplainloaded \else
+  \checkfont{msam10}
+  \iffontfound
+    \IfFileExists{amssymb.sty}
+      {\typeout{^^JFound AMS Symbol fonts on the system, using the
+                'amssymb' package.^^J}%
+       \usepackage{amssymb}%
+       \let\le=\leqslant  \let\leq=\leqslant
+       \let\ge=\geqslant  \let\geq=\geqslant
+      }{}
+  \fi
+\fi
+
+% See if the author has the AMS 'amsbsy' package installed: If they have,
+% use it to provide better bold math support (with \boldsymbol).
+
+\ifCUPmtlplainloaded \else
+  \IfFileExists{amsbsy.sty}
+    {\typeout{^^JFound the 'amsbsy' package on the system, using it.^^J}%
+     \usepackage{amsbsy}}
+    {\providecommand\boldsymbol[1]{\mbox{\boldmath $##1$}}}
+\fi
+
+%%% Example macros (some are not used in this sample file) %%%
+
+% For units of measure
+\newcommand\dynpercm{\nobreak\mbox{$\;$dyn\,cm$^{-1}$}}
+\newcommand\cmpermin{\nobreak\mbox{$\;$cm\,min$^{-1}$}}
+
+% Various bold symbols
+\providecommand\bnabla{\boldsymbol{\nabla}}
+\providecommand\bcdot{\boldsymbol{\cdot}}
+\newcommand\biS{\boldsymbol{S}}
+\newcommand\etb{\boldsymbol{\eta}}
+
+% For multiletter symbols
+\newcommand\Real{\mbox{Re}} % cf plain TeX's \Re and Reynolds number
+\newcommand\Imag{\mbox{Im}} % cf plain TeX's \Im
+\newcommand\Rey{\mbox{\textit{Re}}}  % Reynolds number
+\newcommand\Pran{\mbox{\textit{Pr}}} % Prandtl number, cf TeX's \Pr product
+\newcommand\Pen{\mbox{\textit{Pe}}}  % Peclet number
+\newcommand\Ai{\mbox{Ai}}            % Airy function
+\newcommand\Bi{\mbox{Bi}}            % Airy function
+
+% For sans serif characters:
+% The following macros are setup in JPP.cls for sans-serif fonts in text
+% and math.  If you use these macros in your article, the required fonts
+% will be substitued when you article is typeset by the typesetter.
+%
+% \textsfi, \mathsfi   : sans-serif slanted
+% \textsfb, \mathsfb   : sans-serif bold
+% \textsfbi, \mathsfbi : sans-serif bold slanted (doesnt exist in CM fonts)
+%
+% For san-serif roman use \textsf and \mathsf as normal.
+%
+\newcommand\ssC{\mathsf{C}}    % for sans serif C
+\newcommand\sfsP{\mathsfi{P}}  % for sans serif sloping P
+\newcommand\slsQ{\mathsfbi{Q}} % for sans serif bold-sloping Q
+
+% Hat position
+\newcommand\hatp{\skew3\hat{p}}      % p with hat
+\newcommand\hatR{\skew3\hat{R}}      % R with hat
+\newcommand\hatRR{\skew3\hat{\hatR}} % R with 2 hats
+\newcommand\doubletildesigma{\skew2\tilde{\skew2\tilde{\Sigma}}}
+%       italic Sigma with double tilde
+
+% array strut to make delimiters come out right size both ends
+\newsavebox{\astrutbox}
+\sbox{\astrutbox}{\rule[-5pt]{0pt}{20pt}}
+\newcommand{\astrut}{\usebox{\astrutbox}}
+
+\newcommand\GaPQ{\ensuremath{G_a(P,Q)}}
+\newcommand\GsPQ{\ensuremath{G_s(P,Q)}}
+\newcommand\p{\ensuremath{\partial}}
+\newcommand\tti{\ensuremath{\rightarrow\infty}}
+\newcommand\kgd{\ensuremath{k\gamma d}}
+\newcommand\shalf{\ensuremath{{\scriptstyle\frac{1}{2}}}}
+\newcommand\sh{\ensuremath{^{\shalf}}}
+\newcommand\smh{\ensuremath{^{-\shalf}}}
+\newcommand\squart{\ensuremath{{\textstyle\frac{1}{4}}}}
+\newcommand\thalf{\ensuremath{{\textstyle\frac{1}{2}}}}
+\newcommand\Gat{\ensuremath{\widetilde{G_a}}}
+\newcommand\ttz{\ensuremath{\rightarrow 0}}
+\newcommand\ndq{\ensuremath{\frac{\mbox{$\partial$}}{\mbox{$\partial$} n_q}}}
+\newcommand\sumjm{\ensuremath{\sum_{j=1}^{M}}}
+\newcommand\pvi{\ensuremath{\int_0^{\infty}%
+  \mskip \ifCUPmtlplainloaded -30mu\else -33mu\fi -\quad}}
+
+\newcommand\etal{\mbox{\textit{et al.}}}
+\newcommand\etc{etc.\ }
+\newcommand\eg{e.g.\ }
+
+\newtheorem{lemma}{Lemma}
+\newtheorem{corollary}{Corollary}
+\newtheorem{statement}{Statement}
+
+\title[Generalized universal instability]{Generalized universal instability: Transient linear amplification and subcritical turbulence}
+
+%\author[A. N. Other, H.-C. Smith and J. Q. Public]%
+%{A\ls L\ls A\ls N\ns N.\ns O\ls T\ls H\ls E\ls R$^1$%
+%  \thanks{Email address for correspondence: plaeditorial@cambridge.org},\ns
+%H.\ls-\ls C.\ns S\ls M\ls I\ls T\ls H$^1$\break
+%\and J.\ns Q.\ns P\ls U\ls B\ls L\ls I\ls C$^2$}
+%
+%% NOTE: A full address must be provided: department, university/institution, town/city, zipcode/postcode, country.
+%\affiliation{$^1$Plasma Physics Laboratory, University of America,
+%Somewhere, IN 12345, USA\\[\affilskip]
+%$^2$Department of Physics, University of
+%Camford, Academic Street, Camford CF3 5QL, UK}
+
+\author[Matt Landreman, Gabriel G. Plunk and William Dorland]
+{M\ls A\ls T\ls T\ns   L\ls A\ls N\ls D\ls R\ls E\ls M\ls A\ls N$^1$
+  \thanks{Email address for correspondence: mattland@umd.edu},\ns
+G\ls A\ls B\ls R\ls I\ls E\ls L\ns   G.\ns  P\ls L\ls U\ls N\ls K$^2$\break
+\and W\ls I\ls L\ls L\ls I\ls A\ls M\ns   D\ls O\ls R\ls L\ls A\ls N\ls D$^1$}
+
+\affiliation{$^1$Institute for Research in Electronics and Applied Physics, University of Maryland, College Park, MD, 20742, USA\\[\affilskip]
+$^2$Max Planck Institute for Plasma Physics, Max-Planck/Princeton Research Center for Plasma Physics,
+17491 Greifswald, Germany}
+
+\pubyear{2015}
+\volume{??}
+\pagerange{??}
+% Do not enter received and revised dates. These will be entered by the editorial office.
+\date{?; revised ?; accepted ?. - To be entered by editorial office}
+%\setcounter{page}{1}
+\begin{document}
+
+%\documentclass[12pt,superscriptaddress]{revtex4}
+%\usepackage{latexsym, color, graphicx}
+%\usepackage{amssymb}
+%\usepackage{amsmath}
+
+%\newcommand{\red}[1]{\textcolor{red}{#1}}
+%\newcommand{\vect}[1]{\mbox{\boldmath $#1$}}
+\newcommand{\vect}[1]{\boldsymbol{$#1$}}
+%\newcommand{\erf}{\mathrm{erf}}
+%\newcommand{\PS}{Pfirsch-Schl\"{u}ter~}
+%\newcommand{\ExB}{\vect{E}\times\vect{B}}
+%\newcommand{\Cnl}{C_{\mathrm{nl}a}}
+%\newcommand{\vma}{\vect{v}_{\mathrm{m}a}}
+%\newcommand{\vm}{\vect{v}_{\mathrm{m}}}
+%\newcommand{\vda}{\vect{v}_{\mathrm{d}a}}
+%\newcommand{\vE}{\vect{v}_{E}}
+%\newcommand{\vdao}{\vect{v}_{\mathrm{d}a0}}
+%\newcommand{\vdo}{\vect{v}_{\mathrm{d}0}}
+%\newcommand{\vEo}{\vect{v}_{E0}}
+%\newcommand{\vpar}{v_{||}}
+%\newcommand{\vperp}{v_{\bot}}
+%\newcommand{\rhop}{\rho_{\theta}}
+%\newcommand{\rT}{r_{T}}
+%\newcommand{\Ti}{T_{\mathrm{i}}}
+%\newcommand{\rTi}{r_{T\mathrm{i}}}
+%\newcommand{\rTe}{r_{T\mathrm{e}}}
+%\newcommand{\reta}{r_{\eta}}
+%\newcommand{\rn}{r_{n}}
+%\newcommand{\rperp}{r_\bot}
+%\newcommand{\fM}{f_{\mathrm{M}}}
+%\newcommand{\fMa}{f_{\mathrm{M}a}}
+%\newcommand{\fMi}{f_{\mathrm{M}i}}
+%\newcommand{\fMe}{f_{\mathrm{Me}}}
+%\newcommand{\fe}{f_{\mathrm{e}}}
+%\newcommand{\Te}{T_{\mathrm{e}}}
+%\newcommand{\vth}{v_{\mathrm{th}}}
+%\newcommand{\nuii}{\nu_{ii}}
+%\newcommand{\nuPrime}{\hat{\nu}}
+%\newcommand{\gradpar}{\nabla_{||}}
+%\newcommand{\kpar}{k_{||}}
+%\newcommand{\psia}{\psi_0}
+%\newcommand{\psiaHat}{\hat{\psi}_0}
 \newcommand{\omegase}{\omega_{*}}
@@
 \newcommand{\energy}{\mathcal{E}}
-\newcommand{\sgn}{\mathrm{sgn}}
-\newcommand{\changed}[1]{\textcolor{red}{#1}}
+%\newcommand{\sgn}{\mathrm{sgn}}
+\newcommand{\sgn}{\mbox{sgn}}
+%%\newcommand{\changed}[1]{\textcolor{red}{#1}}
 %\newcommand{\changed}[1]{#1}
 \newcommand{\transpose}{\mathrm{T}}
-
-\begin{document}
-
-\title{Generalized universal instability: Transient linear amplification and subcritical turbulence}
 
@@
 
-\author{Matt Landreman}
-\email[]{mattland@umd.edu}
-\affiliation{Institute for Research in Electronics and Applied Physics, University of Maryland, College Park, MD, 20742, USA}
-\author{Gabriel G. Plunk}
-\affiliation{Max Planck Institute for Plasma Physics, EURATOM Association and Max-Planck/Princeton Research Center for Plasma Physics}
-\author{William Dorland}
-\affiliation{Institute for Research in Electronics and Applied Physics, University of Maryland, College Park, MD, 20742, USA}
-%\homepage[]{Your web page}
-%\thanks{}
-%\author{co-authors}
-
-% Collaboration name, if desired (requires use of superscriptaddress option in \documentclass).
-% \noaffiliation is required (may also be used with the \author command).
-%\collaboration{}
-%\noaffiliation
-
-\date{\today}
+\maketitle
 
@@
 
-\pacs{}% insert suggested PACS numbers in braces on next line
-
-\maketitle %\maketitle must follow title, authors, abstract and \pacs
+\begin{PACS}
+???
+\end{PACS}
 
@@
 eigenmode analysis can in practice give misleading predictions
-about the stability of various systems \cite{TrefethenEmbree}.
+about the stability of various systems \citep{TrefethenEmbree}.
 In particular, turbulence may be sustained in a number of fluid systems
@@
 a phenomenon that may be called ``subcritical turbulence''.
-The first systems known to display this behavior were certain sheared flow patterns in
+The first systems known to display this behaviour were certain sheared flow patterns in
 neutral fluids, such as plane Poiseuille or plane Couette flow
-\cite{Tillmark,TrefethenSubcritical,Grossman}.
+\citep{Tillmark,TrefethenSubcritical,Grossman}.
 Later, sustained turbulence was seen
 in fluid simulations of several plasma systems with no absolute linear instabilities
-\cite{Waltz, Scott1, Scott2, Drake}.
+\citep{Waltz, Scott1, Scott2, Drake}.
 More recently,
 turbulence in the absence of linear instability has been seen in kinetic plasma simulations involving
-sheared equilibrium flow \cite{EdmundPRL, BarnesRotation}.
+sheared equilibrium flow \citep{EdmundPRL, BarnesRotation}.
 These examples demonstrate that linear eigenmode analysis can give
-deceiving predictions for nonlinear behavior,
+deceiving predictions for nonlinear behaviour,
 a conclusion supported by other studies of plasma turbulence in which
@@
 the fluctuations in saturated nonlinear turbulence are different in character from the linear
-eigenmodes \cite{Biskamp,Friedman1,Friedman2}.
+eigenmodes \citep{Biskamp,Friedman1,Friedman2}.
 
@@
 appears to be important
-\cite{TrefethenSubcritical, FarrellIoannou, Grossman, DelSoleSurvey, TrefethenEmbree, Schmid, Newton, Alex}.
+\citep{TrefethenSubcritical, FarrellIoannou, Grossman, DelSoleSurvey, TrefethenEmbree, Schmid, Newton, Alex}.
 Also known as non-modal amplification, transient amplification is a period of growth before
@@
 Analysis of transient amplification has also been applied to plasma systems in which
-linear instability exists but transient linear behavior may be
-more important \cite{Camargo, Camporeale2010, Friedman2, SquirePRL, SquireApJ}.
+linear instability exists but transient linear behaviour may be
+more important \citep{Camargo, Camporeale2010, Friedman2, SquirePRL, SquireApJ}.
 
@@
 without instability for the case of the weakly-collisional
-plasma drift wave driven by a density gradient\cite{Galeev,Krall3},
+plasma drift wave driven by a density gradient \citep{Galeev,Krall3},
 neglecting toroidal effects.
@@
 where $k_y$ is the perpendicular wavenumber and $\rho_i$ is the ion gyroradius
-\cite{Ross, Tsang, Antonsen}.
+\citep{Ross, Tsang, Antonsen}.
 However, instability can persist at $k_y \rho_i \ge 1$ when the collisionality is
-very low \cite{usUniversalInstability}.
-When both magnetic shear and modest collisions are included \cite{usUniversalInstability},
+very low \citep{usUniversalInstability}.
+When both magnetic shear and modest collisions are included,
 the mode becomes linearly stable at all $k_y \rho_i$.
@@
 universal instability, but also for the slab ion-temperature-gradient (ITG) instability.
-Specifically, in \cite{Plunk2014}, considering the slab mode by setting
+Specifically, in \citet{Plunk2014}, considering the slab mode by setting
 curvature drift effects $\omega_d$ to 0, the small-shear limit of (19) is different from
@@
 
-Fluid versions of this drift wave system were investigated in \cite{Scott1,Scott2,Drake}.
-%using a model with three fluctuating quantities (density, electrostatic potential, and parallel velocity).
+Fluid versions of this drift wave system were investigated by \citet{Scott1,Scott2}
+and \citet{Drake}.
 In these models, all linear modes were damped, yet nonlinearly, turbulence could be sustained.
-In \cite{Drake} it was pointed out that an important secondary
+In \citet{Drake} it was pointed out that an important secondary
 instability in this system is a drift wave rotated by $\pi/2$ in the plane
@@
 Several authors have also investigated transient linear effects in fluid drift wave models
-\cite{Camargo, Friedman1, Friedman2}. However, these studies did not
+\citep{Camargo, Friedman1, Friedman2}. However, these studies did not
 include magnetic shear, which we will find to have a very strong effect on
-some aspects of the transient behavior.
+some aspects of the transient behaviour.
 
@@
 extending previous work by considering a kinetic description and exploring the role of transient linear amplification.
-For these studies, the initial-value gyrokinetic code gs2 \cite{gs2} will be used
+For these studies, the initial-value gyrokinetic code gs2 \citep{gs2} will be used
 for numerical computations.
@@
 Second, in section \ref{sec:turbulence} we demonstrate that universal-mode turbulence without linear
-instability, seen previously in fluid models in \cite{Scott1,Scott2,Drake},
+instability, seen previously in fluid models  \citep{Scott1,Scott2,Drake},
 is also possible in the gyrokinetic model.
@@
 transient linear amplification and nonlinear turbulence in this system,
-using similar arguments to the proof for hydrodynamics in \cite{DelSoleNecessity}
+using similar arguments to the proof for hydrodynamics in \citet{DelSoleNecessity}
 and arriving at similar conclusions.
@@
 In this geometry, we consider the nonlinear electrostatic
-gyrokinetic equation \cite{FriemanChen},
+gyrokinetic equation \citep{FriemanChen},
 dropping toroidal effects:
@@
 $T_s$ is the species temperature,
-\changed{$\eta_s=(d \ln T_s/dx)/(d \ln n/dx)$,}
+$\eta_s=(d \ln T_s/dx)/(d \ln n/dx)$,
 $C_s$ is the collision operator,
@@
 The distinction between $z$ and $Z_s$ may be neglected.
+We consider a flux tube domain, in which the fluctuating quantities can be taken
+to be periodic in $x$ and $y$.
 Expanding the fluctuating quantities in spatial Fourier modes as $\Phi = \sum_{k_x,k_y} \Phi_{k_x,k_y}(t,z) \exp(i k_x x + i k_y y)$
@@
 One important property of the system (\ref{eq:gke})-(\ref{eq:qn})
-is the free energy conservation equation \cite{Howes, AbelReview}
-obtained from the $\sum_s \int d^3R_s \int d^3v (T_s h_s/f_{Ms}) (\ldots)$ moment of (\ref{eq:gke}):
+is the free energy conservation equation \citep{Sugama1996, Howes, AbelReview}
+obtained by applying the operation $\sum_s \int d^3R_s \int d^3v (T_s h_s/f_{Ms}) (\ldots)$ to (\ref{eq:gke}):
 \begin{equation}
@@
 = \int d^3 r \int d^3v \left< \ldots\right>_{\vect{r}}
-$ are used.)
+$ are used.
+The $v_{||} \partial h_s/\partial z$ term in (\ref{eq:gke}) has been integrated by parts;
+as detailed in \citet[pp. 66]{BanonNavarroThesis}, the associated boundary term vanishes for a finite $z$ domain
+using a twist-and-shift parallel boundary condition \citep{Beer1995},
+taking the fluctuation amplitude to vanish at the ends of each set of linked domains.)
 The first term on the right-hand side of (\ref{eq:freeEnergySources})
@@
 no longer recognizable as the unsheared slab modes, and indeed are absolutely stable
-for $k_y\rho_i \ll 1$ \cite{Ross,Tsang}.  One might reasonably expect weakly sheared systems with
+for $k_y\rho_i \ll 1$ \citep{Ross,Tsang}.  One might reasonably expect weakly sheared systems with
 stable global eigenmodes to exhibit transient linear amplification, with the number of $e$-foldings
@@
 Collisions are also included to eliminate absolute instability for $k_y \rho_i \ge 0.7$,
-as discussed in \cite{usUniversalInstability},
-using the collision operators detailed in Ref. \cite{Abel}.
+as discussed in \citet{usUniversalInstability},
+using the collision operators detailed in \citet{Abel}.
 As transient amplification generally depends on the choice
@@
 
-Results for the three norms shown are quite similar,
-though the curve for the $W$ norm appears slightly higher than the other curves
-due to less decay in the initial period $t v_i / L_n <20$.
-Indeed, we find
-all results described or plotted for the remainder of this paper are quite similar
-for all of these norms.
-Henceforth we restrict our attention to the $W$ norm since it is well motivated
-by the conservation equation (\ref{eq:dWdt}).
-
-For the calculation in figure \ref{fig:transientAmplification}, the initial condition used
-is the standard ``noise'' initial condition of gs2:
-$g_s = h_s - \left< \Phi\right>_{\vect{R}_s} q_s f_{Ms} / T_s$ is set to a Maxwellian velocity distribution
-times a random complex number at each grid point in $z$.
-(The real and imaginary parts of these complex numbers are each taken from a uniform distribution on $[-1,1]$.)
-While techniques exist to calculate the optimal perturbation
-that gives maximum amplification \cite{FarrellIoannou, SquirePRL, SquireApJ}, such techniques
-are numerically challenging for this kinetic model due to the high dimensionality
-of the problem.
-The noise initial condition is a convenient shortcut, ensuring energy is likely
-to be given to a growing perturbation if one exists.
-It has been shown in \cite{Camargo, SquireApJ} that for studies of transient linear
-amplification, random initial conditions are a reasonable proxy
+For the calculation in figure \ref{fig:transientAmplification}, the
+initial condition used is the standard ``noise'' initial condition of
+gs2: $g_s = h_s - \left< \Phi\right>_{\vect{R}_s} q_s f_{Ms} / T_s$ is
+set to a Maxwellian velocity distribution times a random complex
+number at each grid point in $z$.  (The real and imaginary parts of
+these complex numbers are each taken from a uniform distribution on
+$[-1,1]$.)  While techniques exist to calculate the optimal
+perturbation that gives maximum amplification \citep{FarrellIoannou,
+  SquirePRL, SquireApJ}, such techniques are numerically challenging
+for this kinetic model due to the high dimensionality of the problem.
+The noise initial condition is a convenient shortcut, ensuring energy
+is likely to be given to a growing perturbation if one exists.  It has
+been shown in \citet{Camargo} and \citet{SquireApJ} that for studies of transient
+linear amplification, random initial conditions are a reasonable proxy
 for a directly calculated optimal initial condition.
-
-\begin{figure}[h!]
+Figure \ref{fig:transientAmplification}.b shows that 
+when the calculation is repeated many times, the dynamics of the energy in figure \ref{fig:transientAmplification}.a are typical.
+
+Results for the three norms shown in figure
+\ref{fig:transientAmplification}.a are quite similar, though the curve
+for the $W$ norm appears slightly higher than the other curves due to
+less decay in the initial period $t v_i / L_n <20$.  Indeed, we find
+all results described or plotted for the remainder of this paper are
+quite similar for all of these norms, 
+using the locally-Maxwellian
+initial perturbations described above.  However, there are certain
+non-Maxwellian initial perturbations that can cause arbitrarily fast
+growth in the electric field energy norms but not in the $W$ norm \citep{GregIan}.
+Furthermore, it is $W$ that is conserved by the nonlinearities (recall
+the conservation equation (\ref{eq:dWdt})) and thus the $W$ norm that
+plays a role in determining bounds on the possibility of
+nonlinearly-sustained turbulence, as we will show in section
+\ref{sec:proof}.  For these reasons, we henceforth restrict our attention
+to the $W$ norm.
+
+\begin{figure}
 % Figure generated with m20141119_01_plotGS2TransientAmplificationForPaper_figureA.m
-\includegraphics[width=3in]{m20141119_01_plotGS2TransientAmplificationForPaper_figureA_a.pdf}
+\centerline{
+%\includegraphics[width=3in]{m20141119_01_plotGS2TransientAmplificationForPaper_figureA_a.pdf}
+\includegraphics[width=3in]{fig1a.eps}
+%\includegraphics[width=3in]{m20150305_02_plotManyIndntclGS2TransntAmplifctnRuns.pdf}
+\includegraphics[width=3in]{fig1b.eps}
+}
 % Figure generated with m20141121_01_plotGS2TransientAmplificationForPaper.m
-\includegraphics[width=3in]{m20141121_01_plotGS2TransientAmplificationForPaper_figuresBC_b.pdf}
-\includegraphics[width=3in]{m20141121_01_plotGS2TransientAmplificationForPaper_figuresBC_c.pdf}
-\caption{(Color online) (a) Linear evolution for $L_s/L_n=300$, $\eta_i = \eta_e = 0$,
+\centerline{
+%\includegraphics[width=3in]{m20141121_01_plotGS2TransientAmplificationForPaper_figuresBC_c.pdf}
+\includegraphics[width=3in]{fig1c.eps}
+%\includegraphics[width=3in]{m20141121_01_plotGS2TransientAmplificationForPaper_figuresBC_d.pdf}
+\includegraphics[width=3in]{fig1d.eps}
+}
+\caption{(Colour online) (a) Linear evolution for $L_s/L_n=300$, $\eta_i = \eta_e = 0$,
 $\nuee = 0.05 v_i/L_n$, and $k_y \rho_i = 1$, illustrating transient amplification in three norms.
 The initial conditions are detailed in the main text.
-(b) The amount of transient amplification may be defined relative to $t=0$ or relative
+(b) Repeating the calculation 50 times, the results are always qualitatively similar.
+(c) The amount of transient amplification may be defined relative to $t=0$ or relative
 to the minimum that occurs shortly thereafter.
-(c) The potential structure along a field line for this example.
+(d) The potential structure along a field line for this example.
 \label{fig:transientAmplification}}
@@
 $k_{\perp}$ at each time for the transient fluctuation by
-\changed{
 \begin{equation}
@@
 \end{equation}
-}
 Similarly, we define an effective $|k_{||}|$ at each time for the transient structure
@@
 The resulting growth rate 
-\changed{
-and real frequency can then be compared to the behavior of the finite-shear
+and real frequency can then be compared to the behaviour of the finite-shear
 transient structure.
@@
 instability at the average wavenumbers $\left< k_{\perp} \right>$ and $\left< \left| k_{||} \right| \right>$.
-}
-
-\begin{figure}[h!]
-\includegraphics[width=4in]{m20150222_01_plotGS2TransientAmplificationVs0ShearEigenmode.pdf}
-\caption{(Color online)
-\changed{
+
+\begin{figure}
+%\centerline{\includegraphics[width=4in]{m20150222_01_plotGS2TransientAmplificationVs0ShearEigenmode.pdf}}
+\centerline{\includegraphics[width=4in]{fig2.eps}}
+\caption{(Colour online)
 Comparison of the finite-shear transient from Figure \ref{fig:transientAmplification} to a
@@
 (The amplitude and phase of the eigenmode in (b) are arbitrary.)
-%from (\ref{eq:expectedKperp})-(\ref{eq:expectedKpar}).
-}
 \label{fig:comparingTransientTo0Shear_single}}
@@
 
-\changed{
 To demonstrate this principle across a range of parameters,
@@
 correspondence between the finite-shear transients and zero-shear eigenmodes
-is maintained throughout the evolution of the transient and for other
-values of magetic shear.}
-%This comparison is shown in figure \ref{fig:comparingTransientTo0Shear}.
+is maintained throughout the evolution of the transient, for a range of
+values of magetic shear.
 In this figure, we display 25 independent simulations of linear transients,
@@
 For all simulations shown, $k_y \rho_i = 2$ and $\nuee = 0.05 v_i/L_n$.
-In each case, there is very close agreement between the instantaneous growth rate of the transient
+Both $\left<|k_{||}|\right>$ and $\left< k_\perp \right>$ increase with time during the transient
+growth, corresponding to movement upward and rightward in figure \ref{fig:localDispersionRelation}.a.
+As a result, the effective zero-shear growth rate decreases over time.
+For all transients we examined, there is very close agreement between the instantaneous growth rate of the transient
 and the `predicted' growth rate from the zero-shear dispersion relation.
@@
 
-\begin{figure}[h!]
-% Data generated on Edison in /global/scratch2/sd/landrema/20150109-01-coderunner-edison-0shearUniversalInstability/vnewk_e_0p05
+\begin{figure}
+% Data for finite-shear transients were generated on Edison in
+% /global/scratch2/sd/landrema/20141121-01-coderunner-edison-transientAmplificationScans/aky2/
+%
+% Data for zero-shear dispersion relation were generated on Edison in 
+% /global/scratch2/sd/landrema/20150109-01-coderunner-edison-0shearUniversalInstability/vnewk_e_0p05
+%
 % Figure generated with m20150109_04_cmprngTransientGrowthRateTo0ShearForPaper.m
-\includegraphics[width=4in]{m20150109_04_cmprngTransientGrowthRateTo0ShearForPaper.pdf}
-\caption{(Color online)
+%\centerline{\includegraphics[width=4in]{m20150109_04_cmprngTransientGrowthRateTo0ShearForPaper.pdf}}
+\centerline{\includegraphics[width=4in]{fig3.eps}}
+\caption{(Colour online)
 Although a change from zero to small finite magnetic shear causes a fundamental change in the eigenvalues,
@@
 are driven by the combination of resonant electrons and the density gradient.
+
+In contrast to the agreement shown in figure
+\ref{fig:comparingTransientTo0Shear} for the amplification phase
+following the maximum growth rate, we do not find such close agreement
+between the zero-shear dispersion relation and the finite-shear
+dynamics for very early times, before the maximum growth rate occurs.
+In this initial period, a single dominant wave packet has not yet
+overtaken other fluctuations present in the initial random noise.  We
+also do not see such close agreement between the zero-shear and
+finite-shear behaviour during the decay phase of the latter, i.e. at
+large times.  The finite-shear decay rate as $t \to \infty$
+corresponds to an eigenvalue, and it was established in previous work
+\citep{Ross,Tsang,Antonsen} that the finite-shear eigenvalues
+are not continuously related to the zero-shear eigenvalues.  In an
+eigenmode, points at all $z$ have a coherent time-dependence.  Thus,
+this damping rate is determined by a coherent interaction between
+points of different $k_\perp(z)=k_y\sqrt{1+(z/L_s)^2}$, and between
+all $k_{||}$ Fourier components.  This damping rate is thus not well
+approximated by the zero-shear dispersion relation which assumes a
+single $k_\perp$ and single $k_{||}$.
 
@@
 
-\changed{
 While figures \ref{fig:comparingTransientTo0Shear_single}-\ref{fig:comparingTransientTo0Shear}
@@
 of the overall $W(t)$ curve for each transient seen in figures \ref{fig:transientAmplificationVsLsOverLn}.a-b.
-}
 
@@
 (When the collisionality is much smaller than these values,
-absolute linear instability is present \cite{usUniversalInstability}.)
+absolute linear instability is present \citep{usUniversalInstability}.)
 It can be seen that increasing collisionality tends to limit amplification
@@
 for the hydrodynamic case.)
-However, the kinetic problem here is rather different
-from the hydrodynamic case in the sensitivity of the transient amplification
-to the geometric parameter $L_s/L_n$,
-which is not associated with entropy-producing collisions.
-
-\begin{figure}[h!]
+
+\begin{figure}
 % Data computed on Edison in /global/scratch2/sd/landrema/20141121-01-coderunner-edison-transientAmplificationScans
@@
 % m20140723_03_plotTransientAmplificationLsOverLnScanForPaper.m on my laptop.
-\includegraphics[width=3in]{m20141121_03_plotTransientAmplificationLsOverLnScanForPaper_a.pdf}
-\includegraphics[width=3in]{m20141121_03_plotTransientAmplificationLsOverLnScanForPaper_b.pdf}
-%\includegraphics[width=3in]{m20140723_03_plotTransientAmplificationLsOverLnScanForPaper.pdf}
-\caption{(Color online) The (a) number of $e$-foldings of transient amplification and (b)
+\centerline{
+%\includegraphics[width=3in]{m20141121_03_plotTransientAmplificationLsOverLnScanForPaper_a.pdf}
+%\includegraphics[width=3in]{m20141121_03_plotTransientAmplificationLsOverLnScanForPaper_b.pdf}
+\includegraphics[width=3in]{fig4a.eps}
+\includegraphics[width=3in]{fig4b.eps}
+}
+\caption{(Colour online) The (a) number of $e$-foldings of transient amplification and (b)
 duration of the amplification both scale with the magnetic shear length scale $L_s$,
@@
 
-\begin{figure}[h!]
+\begin{figure}
 % Data computed on Edison in /global/scratch2/sd/landrema/20141121-02-coderunner-edison-transientAmplificationKyScans
@@
 % m20140728_01_plotTransientAmplificationKyScans.m on my laptop.
-\includegraphics[width=3in]{m20141122_01_plotTransientAmplificationKyScans.pdf}
-\includegraphics[width=3in]{m20141122_01_plotTransientAmplificationKyScans_times.pdf}
-%\includegraphics[width=3in]{m20140728_01_plotTransientAmplificationKyScans.pdf}
-\caption{(Color online) Transient amplification can be observed
+\centerline{
+%\includegraphics[width=3in]{m20141122_01_plotTransientAmplificationKyScans.pdf}
+%\includegraphics[width=3in]{m20141122_01_plotTransientAmplificationKyScans_times.pdf}
+\includegraphics[width=3in]{fig5a.eps}
+\includegraphics[width=3in]{fig5b.eps}
+}
+\caption{(Colour online) Transient amplification can be observed
 for a range of $k_y$ surrounding $1/\rho_i$.  For this figure, $L_s/L_n=200$.
@@
 \end{figure}
+
+In the context of sheared flows in neutral fluids, a relationship
+between transient growth and eigenvalue sensitivity has been formally
+proved using the notion of a pseudospectrum
+\citep{ReddyHenningson,TrefethenSubcritical,TrefethenEmbree}. The
+$\epsilon$-pseudospectrum of a linear operator can be defined as the
+union of the sets of eigenvalues obtained when the operator is
+perturbed by any other operator of norm $\le \epsilon$, providing a
+precise notion of the sensitivity of eigenvalues.  One can rigorously
+bound the amount of possible transient growth by the extent of the
+pseudospectrum into the unstable half of the complex plane
+\citep{ReddyHenningson}, summarized in equation (12) of \citet{TrefethenSubcritical}.
+However, for the universal instability, it is not clear that changing
+the magnetic shear from 0 to a small finite value has the form of the
+operator perturbation used to define the pseudospectrum.  The reason
+is that the $z$ domain is infinite, and for any nonzero shear, there
+is always some $z$ for which the argument of $J_0$ is changed
+significantly compared to the zero-shear case.  Indeed, all evidence
+suggests that when the shear changes from 0 to any small finite value,
+the eigenvalues (for a range of $k_y$) change from unstable to stable
+discontinuously rather than continuously \citep{Ross,Tsang}.  Thus, it
+may or may not be possible to understand the role of magnetic shear in
+the universal instability using the formal theory of pseudospectra,
+and we leave this careful mathematical analysis for future work.
 
@@
 for random initial conditions.
-(While absolute linear instability was found in \cite{usUniversalInstability} for $L_s/L_n \ge 17$ in the limit of vanishing collisionality, here the finite collisionality is enough to suppress this linear instability.)
+(While absolute linear instability was found in \citet{usUniversalInstability} for $L_s/L_n \ge 17$ in the limit of vanishing collisionality, here the finite collisionality is enough to suppress this linear instability.)
 Before presenting nonlinear simulations,
-we first illustrate the linear behavior of the system using all the physical and numerical
+we first illustrate the linear behaviour of the system using all the physical and numerical
 parameters that will be used for the nonlinear runs.  Figure \ref{fig:linearBox} shows the time evolution of
@@
 
-\begin{figure}[h!]
+\begin{figure}
 % Dataset is /global/scratch2/sd/landrema/20141112-01-coderunner-edison-subcritical-box-lowLsOverLn/v/id_59
@@
 % m20140804_03_plotPhi2VsTimeForPaper_linearBox.m on my laptop
-\includegraphics[width=6.5in]{m20141126_01_plotWVsTimeForPaper_linearBox.pdf}
-%\includegraphics[width=6.5in]{m20140804_03_plotPhi2VsTimeForPaper_linearBox.pdf}
-\caption{(Color online)
+%\centerline{\includegraphics[width=6.5in]{m20141126_01_plotWVsTimeForPaper_linearBox.pdf}}
+\centerline{\includegraphics[width=6in]{fig6.eps}}
+\caption{(Colour online)
 Linear dynamics for the parameters used in the
 nonlinear simulations of figure \ref{fig:amplitudeDependence}.
-The large-$t$ behavior demonstrates there is no absolute linear instability.
+The large-$t$ behaviour demonstrates there is no absolute linear instability.
 A small amount of transient amplification can be seen.
@@
 temperature gradient to bring about linear instability, and then to reduce the temperature gradient
-once the turbulence reaches a sufficient amplitude \cite{Drake}.
+once the turbulence reaches a sufficient amplitude \citep{Drake}.
 A third effective method, the one used in figure \ref{fig:amplitudeDependence},
@@
 is minimized.
-
-\begin{figure}[h!]
+(In both simulations shown in figure  \ref{fig:amplitudeDependence}, the $k_y \rho_i=0$ Fourier component is initialized to 0,
+but this component (and the $k_y \rho_i=\pm 0.2$ components) grow up quickly for the following reason. Since observable quantities are real, there is a large $k_y \rho_i=-0.1$
+Fourier component present in the initial condition which is the complex conjugate of the large $k_y \rho_i=+0.1$ component shown,
+and these two components interact nonlinearly to drive components at the sum and difference wave numbers.)
+
+\begin{figure}
 % Datasets are in /global/scratch2/sd/landrema/20141112-01-coderunner-edison-subcritical-box-lowLsOverLn
@@
 % m20140804_05_plotPhi2VsTimeForPaper_amplitudeDependence.m on my laptop.
-\includegraphics[width=6.5in]{m20141125_02_WVsTimeForPaper_amplitudeDependence.pdf}
-%\includegraphics[width=6.5in]{m20140804_05_plotPhi2VsTimeForPaper_amplitudeDependence.pdf}
-\caption{(Color online)
+%\centerline{\includegraphics[width=6.5in]{m20141125_02_WVsTimeForPaper_amplitudeDependence.pdf}}
+\centerline{\includegraphics[width=6.5in]{fig7.eps}}
+\caption{(Colour online)
 Simulations for exactly the same physical and numerical parameters as the decaying
@@
 
-\begin{figure}[h!]
+\begin{figure}
 % Simulations in /global/scratch2/sd/landrema/20141112-01-coderunner-edison-subcritical-box-lowLsOverLn
@@
 % m20140807_02_plotPhi2VsTimeForPaper_linearRestart.m on my laptop.
-\includegraphics[width=6.5in]{m20141126_02_plotWVsTimeForPaper_linearRestart.pdf}
-%\includegraphics[width=6.5in]{m20140807_02_plotPhi2VsTimeForPaper_linearRestart.pdf}
-\caption{(Color online)
+%\centerline{\includegraphics[width=6.5in]{m20141126_02_plotWVsTimeForPaper_linearRestart.pdf}}
+\centerline{\includegraphics[width=6.5in]{fig8.eps}}
+\caption{(Colour online)
 Demonstration of the concepts in section \ref{sec:proof}.
@@
 \left(T / 500 \mathrm{\;eV}\right)^{3/2}
+\times
 \left(B/2\mathrm{\;T}\right)^{-2} \left( L_n / 1 \mathrm{\;cm}\right)^{-1}$.
@@
 
-\begin{figure}[h!]
+\begin{figure}
 % Runs on edison in
@@
 % m20140807_01_plotFluxesVsTimeForPaper.m on my laptop.
-\includegraphics[width=4in]{m20141210_01_plotGs2FluxesVsTimeForPaper.pdf}
-%\includegraphics[width=3in]{m20140807_01_plotFluxesVsTimeForPaper.pdf}
-%\includegraphics[width=3in]{fluxes.png}
-\caption{(Color online)
+%\centerline{\includegraphics[width=4in]{m20141210_01_plotGs2FluxesVsTimeForPaper.pdf}}
+\centerline{\includegraphics[width=4in]{fig9.eps}}
+\caption{(Colour online)
 Fluxes for the simulation of Figure (\ref{fig:amplitudeDependence}).a with  $\nuee L_n / v_i = 0.05$ and
@@
 
-\changed{
-As mentioned earlier, for certain values of shear, collisionality, and initial perturbation amplitude,
-a third type of behavior was sometimes observed besides the two outcomes shown in figure \ref{fig:amplitudeDependence}:
-turbulence resembling figure  \ref{fig:amplitudeDependence}.a can persist for a finite time but eventually decay linearly
-as in figure  \ref{fig:amplitudeDependence}.b.
-This finite-lifetime plasma turbulence is reminiscent of turbulent puffs in neutral (Navier-Stokes) pipe flow.
-In neutral pipe flow it has been found \cite{Hof2006, Tel} that in fact
-the lifetime of turbulence puffs does not become infinite at a critical value of the relevant physical parameter (Reynolds number) but rather
-the lifetime varies with Reynolds number exponentially, so the turbulence lifetime can become longer than any reasonable observation
-time.
-An interesting open question (which we leave for future research) is whether a similar scaling
-of the turbulence lifetime may be present in the plasma system we consider here.
-}
+As mentioned earlier, for certain values of shear, collisionality, and
+initial perturbation amplitude, a third type of behaviour was sometimes
+observed besides the two outcomes shown in figure
+\ref{fig:amplitudeDependence}: turbulence resembling figure
+\ref{fig:amplitudeDependence}.a can persist for a finite time but
+eventually decay linearly as in figure
+\ref{fig:amplitudeDependence}.b.  This finite-lifetime plasma
+turbulence is reminiscent of turbulent puffs in neutral
+(Navier-Stokes) pipe flow.  In neutral pipe flow it has been found
+\citep{Hof2006, Tel} that in fact the lifetime of turbulence puffs does
+not become infinite at a critical value of the relevant physical
+parameter (Reynolds number) but rather the lifetime varies with
+Reynolds number exponentially, so the turbulence lifetime can become
+longer than any reasonable observation time.  An interesting open
+question (which we leave for future research) is whether a similar
+scaling of the turbulence lifetime may be present in the plasma system
+we consider here.
 
@@
 
-It seems plausible that the previous two sections are related,
-that a connection exists between transient linear amplification and sustained nonlinear turbulence.
-Let us now prove this relationship rigorously.
-Here we present two such arguments,
-both of which yield the conclusion that non-modal amplification is a necessary condition
-for sustained turbulence.  One argument for the case of hydrodynamics
-was given in \cite{DelSoleNecessity} and section 4 of \cite{DelSoleSurvey},
-and below we adapt this reasoning to gyrokinetics.  Before we do so, we give an alternative
-argument that relies less on linear algebra.
-For this discussion, it is useful to define a ``2D mode'' to be any structure with given $k_x$ and $k_y$,
-but with no restriction on the $z$ dependence (so the structure need not be an eigenmode).
-
-{\it Statement 1:} Consider a turbulent solution of the nonlinear gyrokinetic system (\ref{eq:gke})-(\ref{eq:qn}),
+It seems plausible that the previous two sections are related, that a
+connection exists between transient linear amplification and sustained
+nonlinear turbulence.  Let us now prove this relationship rigorously.
+Here we present two such arguments, both of which yield the conclusion
+that non-modal amplification is a necessary condition for sustained
+turbulence.  One argument for the case of hydrodynamics was given in
+\citet{DelSoleNecessity} and section 4 of \citet{DelSoleSurvey}, and
+below we adapt this reasoning to gyrokinetics.  Before we do so, we
+give an alternative argument that relies less on linear algebra.  For
+this discussion, it is useful to define a ``2D mode'' to be any
+structure with given $k_x$ and $k_y$, but with no restriction on the
+$z$ dependence (so the structure need not be an eigenmode).
+
+\begin{statement}
+Consider a turbulent solution of the nonlinear gyrokinetic system (\ref{eq:gke})-(\ref{eq:qn}),
 i.e. a solution for which $W$ does not decay to zero at large $t$.
-Choose any time $t_0$ at which the free energy satisfies $dW/dt>0$.
+Choose any time $t_0$ at which the free energy satisfies $dW/dt>0$;
+such times must occur in steady turbulence as $W$ fluctuates about its mean value.
 Suppose at $t_0$ we turn off the nonlinear term
@@
 for some time following $t_0$, i.e. there must be transient linear amplification.
-
-{\it Proof:}
+\end{statement}
+
+\begin{proof}
 Recall that (\ref{eq:dWdt}) is a moment of both the nonlinear and linear dynamical equations.
@@
 so this mode must be growing transiently in the $W$ norm.
+\end{proof}
 
@@
 
+We note that while $dW/dt$ is continuous when the nonlinear term is
+turned off, each $dW_{k_x,k_y}/dt$ is generally not. The reason is
+that in order to annihilate the nonlinear term in the derivation of
+(\ref{eq:dWdt})-(\ref{eq:freeEnergySources}), it is necessary to
+integrate over $x$ and $y$, i.e. sum over $k_x$ and $k_y$. Thus, while
+(\ref{eq:dWdt}) is identical for the nonlinear and linear dynamics,
+the analogous equation for $dW_{k_x,k_y}/dt$ is different for the
+nonlinear and linear dynamics. Removing the nonlinear term therefore
+causes a discontinuous change in $dW_{k_x,k_y}/dt$.
+
 Next, we present a second argument for the necessity of transient amplification.
-This argument is based on linear algebra and closely follows \cite{DelSoleNecessity}.
+This argument is based on linear algebra and closely follows \citet{DelSoleNecessity}.
 For this second approach, we let $\vect{h}$ denote
@@
 
-{\it Statement 2:} Given a set of physical parameters for which
+\begin{statement}
+Given a set of physical parameters for which
 the nonlinear gyrokinetic system (\ref{eq:gke})-(\ref{eq:qn})
@@
 onto $\vect{h}_0$, defined below, has a non-zero average.
-
-{\it Proof:}
+\end{statement}
+
+%{\it Proof:}
+\begin{proof}
 Let $\vect{L}$ denote the time-independent linear operator associated with (\ref{eq:gke})-(\ref{eq:qn})
@@
 proving statement 2.
+\end{proof}
 
@@
 another norm always exists in which transient amplification is impossible.
-The sum of squared eigenmode coefficients is such a norm \cite{FarrellIoannou1993}.)
+The sum of squared eigenmode coefficients is such a norm \citep{FarrellIoannou1993}.)
 
@@
 other models and systems in which an energy conservation equation holds. For example,
-eq (6) in Ref. \cite{Drake} gives the appropriate conservation law for the fluid model
+eq (6) in \citet{Drake} gives the appropriate conservation law for the fluid model
 in that work, and the first line of (6) gives the associated norm to use in place of $W$.
@@
 
-The ``universal mode'' system (slab geometry with a density gradient, electrostatic,
-no sheared equilibrium flow, and weak temperature gradients)
-is known to have a dramatic change in linearly stability
-as the magnetic shear is raised from zero to small finite values
-\cite{Ross, Tsang, Antonsen,usUniversalInstability}.
-As the eigenvalues are so sensitive to magnetic shear,
-we might expect traditional modal analysis to
-give a misleading picture of typical behavior,
-as with other systems that have sensitive eigenvalues \cite{TrefethenEmbree}.
-Indeed, here we have demonstrated that significant
-transient linear amplification is possible in the drift wave system even when
-all eigenmodes are decaying.
-As shown in figure \ref{fig:comparingTransientTo0Shear},
-the transient structures grow at the zero-shear
-growth rate, when evaluated at the dominant $k_{\perp}$ and $k_{||}$ at that instant.
-We have demonstrated that the amount and duration of amplification scale
-with the magnetic shear length $L_s$, so these quantities naturally become
-unbounded in the limit of vanishing shear.
-The non-modal amplification also decreases with increasing collisionality,
-particularly at the shortest perpendicular wavelengths,
-and the amplification tends to be largest at wavelengths near $k_y \rho_i \sim 1-2$.
-In contrast to more famous examples of non-modal amplification
-from hydrodynamics \cite{TrefethenSubcritical}, and to some recent examples from plasma physics \cite{Newton, Alex, BarnesRotation, EdmundPRL},
-the amplification in the plasma drift wave system here
-does not depend on sheared equilibrium flow.
+The ``universal mode'' system (slab geometry with a density gradient,
+electrostatic, no sheared equilibrium flow, and weak temperature
+gradients) is known to have a dramatic change in linearly stability as
+the magnetic shear is raised from zero to small finite values
+\citep{Ross, Tsang, Antonsen,usUniversalInstability}.  As the
+eigenvalues are so sensitive to magnetic shear, we might expect
+traditional modal analysis to give a misleading picture of typical
+behaviour, as with other systems that have sensitive eigenvalues
+\citep{TrefethenEmbree}.  Indeed, here we have demonstrated that
+significant transient linear amplification is possible in the drift
+wave system even when all eigenmodes are decaying.  As shown in figure
+\ref{fig:comparingTransientTo0Shear}, the transient structures grow at
+the zero-shear growth rate, when evaluated at the dominant $k_{\perp}$
+and $k_{||}$ at that instant.  We have demonstrated that the amount
+and duration of amplification scale with the magnetic shear length
+$L_s$, so these quantities naturally become unbounded in the limit of
+vanishing shear.  The non-modal amplification also decreases with
+increasing collisionality, particularly at the shortest perpendicular
+wavelengths, and the amplification tends to be largest at wavelengths
+near $k_y \rho_i \sim 1-2$.  In contrast to more famous examples of
+non-modal amplification from hydrodynamics
+\citep{TrefethenSubcritical}, and to some recent examples from plasma
+physics \citep{Newton, EdmundPRL, BarnesRotation, Alex}, the
+amplification in the plasma drift wave system here does not depend on
+sheared equilibrium flow.
 
@@
 This phenomenon had been seen previously
-in fluid models \cite{Scott1, Scott2, Drake}.
+in fluid models \citep{Scott1, Scott2, Drake}.
 This phenomenon, clearly visible in figure \ref{fig:amplitudeDependence},
-is opposite to the ``Dimits shift'' \cite{Dimits} seen at different physical parameters,
+is opposite to the ``Dimits shift'' \citep{Dimits} seen at different physical parameters,
 in which nonlinear fluctuations
@@
 exclusive. Both are required to support turbulence.
-Some form of secondary instability (as discussed in \cite{Drake}) must be present to transfer energy out of the transiently growing linear
+Some form of secondary instability (as discussed in \citet{Drake}) must be present to transfer energy out of the transiently growing linear
 structures before these structures begin to decay substantially. However, nonlinear processes cannot
@@
 need not preclude the possibility of significant turbulent particle and energy fluxes.
-%and that nonlinear simulations should sometimes be initialized at finite amplitude to examine this possibility.
-
-\begin{acknowledgments}
-This material is based upon work supported by the
-U.S. Department of Energy, Office of Fusion Energy Science,
-under Award Numbers DEFG0293ER54197 and DEFC0208ER54964,
-and the SciDAC Center for the Study of Plasma Microturbulence.
-Computations were performed on the Edison system at
-the National Energy Research Scientific Computing Center, a DOE Office of Science User Facility supported by the Office of Science of the U.S.
-Department of Energy under Contract No.
-DE-AC02-05CH11231.
-Conversations with
-Tom Antonsen, Michael Barnes, Greg Colyer, James Drake, Edmund Highcock, Greg Hammett, and Adil Hassam
-are gratefully acknowledged.
-\end{acknowledgments}
+
+%Acknowledgements should be included at the end of the paper, before
+%the References section or any appendicies, and should be a separate
+%paragraph without a heading. Several anonymous individuals are
+%thanked for contributions to these instructions.
+
+%\begin{acknowledgments}
+This material is based upon work supported by the U.S. Department of
+Energy, Office of Fusion Energy Science, under Award Numbers
+DE-FG02-93ER54197 and DE-FC02-08ER54964, and the SciDAC Center for the
+Study of Plasma Microturbulence.  Computations were performed on the
+Edison system at the National Energy Research Scientific Computing
+Center, a DOE Office of Science User Facility supported by the Office
+of Science of the U.S. Department of Energy under Contract
+No. DE-AC02-05CH11231.  Conversations with Tom Antonsen, Michael
+Barnes, Greg Colyer, James Drake, Edmund Highcock, Greg Hammett, and
+Adil Hassam are gratefully acknowledged.
+%\end{acknowledgments}
 
@@
 Figure \ref{fig:localDispersionRelation} was generated by applying nonlinear root-finding to (\ref{eq:dispersionRelation}),
-verifying the solution agreed with initial-value gs2 simulations for several values of $k_{||}$ and $k_{\perp}$.
+verifying the solution agreed with initial-value gs2 simulations for various values of $k_{||}$ and $k_{\perp}$.
 It can be seen that absolute linear instability exists for sufficiently small values of $\left|k_{||}\right|$
@@
 
-\begin{figure}[h!]
+\begin{figure}
 % Figure generated with m20141214_01_plotUniversalModeDispersionRelationForPaper.m
 %\includegraphics[width=6in]{localLinearDispersionRelation.png}
-\includegraphics[width=6in]{m20141214_01_plotUniversalModeDispersionRelationForPaper.pdf}
-\caption{(Color online) Solution of the linear dispersion relation in the zero-shear (local) limit, i.e. the
+%\centerline{\includegraphics[width=6in]{m20141214_01_plotUniversalModeDispersionRelationForPaper.pdf}}
+\centerline{\includegraphics[width=6in]{fig10.eps}}
+\caption{(Colour online) Solution of the linear dispersion relation in the zero-shear (local) limit, i.e. the
 root of (\ref{eq:dispersionRelation}) with largest imaginary part, for $\eta_i = \eta_e = 0$,
@@
 
+\bibliographystyle{jpp}
+
 \bibliography{universalModeTurbulencePaper}
